---
title: "Session 2 — Data import and exploratory analysis"
subtitle: "Language Analytics in R"
author: "Peter Gilles"
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    code-fold: false
editor: visual
---

## Learning objectives {.scrollable}

By the end of this session you will:

-   Import tabular, textual, and XML language data into R.
-   Use tidyverse workflows for exploratory statistics and visualization.
-   Tokenize and analyze Gutenberg texts with `tidytext`.
-   Extract structured information from LOD XML with `xml2`.
-   Solve format-specific exercises for CSV, text, and XML data.

This session covers three common data formats for language analytics:

1.  **CSV** (tabular data)
2.  **Text** from Project Gutenberg (`tidytext` workflow)
3.  **XML** from the LOD dictionary

Each section ends with dedicated exercises. Explained solutions are provided in `sessions/session-02-solutions.qmd`.

# Setup

First, load the data from the GitHub repository <https://github.com/PeterGilles/Language-Analytics-in-R>. Use either the download function on GitHub or use `git clone` that will allow you to easily update the data with new versions.

This block loads all packages used in this session and defines a helper function for robust file paths.

```{r}
#| label: setup
#| message: false

library(tidyverse)
library(tidytext)
library(gutenbergr)
library(xml2)

resolve_path <- function(filename) {
  if (file.exists(filename)) {
    filename
  } else {
    file.path("sessions", filename)
  }
}
```

What happens:

-   `library(...)` attaches the packages.
-   `resolve_path()` first checks local folder, then falls back to `sessions/`.
-   This avoids path errors when students render from different working directories.

# Part 1 — Working with CSV data

## Read and inspect CSV {.scrollable}

We start with demographic data by commune in `rnrpp-population-commune.csv`.

This block reads the file into a tibble named `pop`.

```{r}
#| label: csv-read
#| message: false

csv_path <- resolve_path("rnrpp-population-commune.csv")
pop <- read_csv(
  csv_path,
  locale = locale(encoding = "ISO-8859-1"),
  show_col_types = FALSE
)

pop
```

What happens:

-   `resolve_path()` finds the file.
-   `read_csv(..., locale = locale(encoding = \"ISO-8859-1\"))` imports text with the correct source encoding.
-   Printing `pop` gives a quick first look at rows and columns.

If your data is in Excel format (`.xlsx`) instead of CSV, use `read_excel()` from package `readxl`:

```{r}
#| label: xlsx-read-example
#| eval: false

# install.packages("readxl")   # run once if needed

xlsx_path <- resolve_path("my-data.xlsx")

excel_data <- readxl::read_excel(
  xlsx_path,
  sheet = 1
)

excel_data
```

What happens:

-   `readxl::read_excel()` reads an Excel worksheet into a tibble.
-   `sheet = 1` selects the first sheet (you can also use a sheet name).
-   The resulting object can be used exactly like `pop` in tidyverse workflows.

This block performs structural inspection (shape and variable names).

```{r}
#| label: csv-inspect

glimpse(pop)
nrow(pop)
ncol(pop)
names(pop)
```

What happens:

-   `glimpse()` shows column types and sample values.
-   `nrow()` and `ncol()` report dataset size.
-   `names()` lists column names for later use in `mutate()` and `summarise()`.

## Basic exploratory statistics {.scrollable}

Create derived variables and inspect the largest communes.

```{r}
#| label: csv-explore

pop_totals <- pop |>
  mutate(
    total_population = FEMMES_MINEURES + HOMMES_MINEURS + FEMMES_MAJEURES + HOMMES_MAJEURS,
    adult_share = (FEMMES_MAJEURES + HOMMES_MAJEURS) / total_population
  )

pop_totals |>
  arrange(desc(total_population)) |>
  select(COMMUNE_NOM, adult_share, total_population) |>
  slice_head(n = 10)
```

What happens:

-   `mutate()` adds two analysis variables: total size and adult proportion.
-   `arrange(desc(...))` ranks communes from largest to smallest.
-   `select()` keeps only the most relevant columns for interpretation.

This block computes one-row summary statistics across all communes.

```{r}
#| label: csv-summary

pop_totals |>
  summarise(
    mean_commune_population = mean(total_population),
    median_commune_population = median(total_population),
    min_commune_population = min(total_population),
    max_commune_population = max(total_population)
  )
```

What happens:

-   `summarise()` reduces many rows to one row.
-   We compute central tendency (`mean`, `median`) and range (`min`, `max`).
-   This gives a fast statistical profile before plotting.

This block visualizes the top 20 communes with a horizontal bar chart.

```{r}
#| label: csv-plot
#| fig-width: 8
#| fig-height: 5

pop_totals |>
  arrange(desc(total_population)) |>
  slice_head(n = 20) |>
  mutate(COMMUNE_NOM = fct_reorder(COMMUNE_NOM, total_population)) |>
  ggplot(aes(x = total_population, y = COMMUNE_NOM)) +
  geom_col(fill = "steelblue") +
  labs(
    title = "Top 20 communes by population",
    x = "Population",
    y = NULL
  )
```

What happens:

-   `slice_head(n = 20)` keeps only the top communes.
-   `fct_reorder()` orders labels by population so the chart is readable.
-   `geom_col()` displays already-aggregated values.

## Exercises

1.  Compute the total number of minors (`FEMMES_MINEURES + HOMMES_MINEURS`) per commune and list the top 15 communes.
2.  Remove the variable `COMMUNE_CODE` from the tibble and rename the French variable names to English.
3.  Filter `pop_totals` to keep only communes with an adult share below 80 %.
4.  Create a histogram of `total_population` and briefly describe its shape.
5.  Export the tibble `pop_totals` to a csv file using `write.csv`.

# Part 2 — Working with text data from `Project Gutenberg` using `tidytext`

We use [`tidytext`](https://www.tidytextmining.com/tidytext) ideas: convert text to **one-token-per-row**, remove stop words, and compare frequencies. Data comes from `Project Gutenberg` using the R package `gutenbergr` (documentation [here](https://docs.ropensci.org/gutenbergr/index.html)) - make sure it is installed and loaded.

## Download and tokenize Gutenberg texts {.scrollable}

This block downloads two books from Project Gutenberg, including title metadata.

```{r}
#| label: gutenberg-download

gutenberg_books <- gutenberg_download(
  c(12, 1342), # Through the Looking-Glass, Pride and Prejudice
  meta_fields = c("author", "title")

)

gutenberg_books
```

What happens:

-   `gutenberg_download()` retrieves cleaned text (header/footer removed).
-   Each row is one line of text, with `gutenberg_id` , `author` and `title`.
-   The table is now ready for tokenization.

This block converts lines into words and removes stop words.

`stop_words` is a built-in lexicon from `tidytext` containing very frequent function words such as articles, pronouns, and prepositions (e.g. `the`, `and`, `of`, `to`). These words are often useful for grammar, but less useful when you want fast topic-oriented exploration of lexical content.

Important caveat: stop-word removal is a modeling choice, not a universal rule. For stylistic analysis, authorship studies, or grammatical research, you may want to keep stop words.

```{r}
#| label: gutenberg-tokenize

tokens <- gutenberg_books |>
  unnest_tokens(word, text) |>
  anti_join(stop_words, by = join_by(word))

tokens
```

What happens:

-   `unnest_tokens(word, text)` turns each line into multiple word rows.
-   Tokens are lowercase by default and punctuation is removed.
-   `anti_join(stop_words, by = join_by(word))` removes tokens that appear in the stop-word list.
-   The result keeps mostly content words, so top-frequency tables are usually more interpretable.

## Explore word frequencies {.scrollable}

This block counts how often each word appears in each title.

```{r}
#| label: gutenberg-count

top_words <- tokens |>
  count(title, word, sort = TRUE)

top_words |>
  group_by(title) |>
  slice_head(n = 15) |>
  ungroup()
```

What happens:

-   `count(title, word)` computes frequencies per book.
-   `sort = TRUE` returns highest counts first.
-   Grouping + `slice_head()` keeps the top words for each book.

This block visualizes top words by title in faceted panels.

```{r}
#| label: gutenberg-plot
#| fig-width: 9
#| fig-height: 5

top_words |>
  group_by(title) |>
  slice_head(n = 12) |>
  ungroup() |>
  mutate(word = fct_reorder(word, n)) |>
  ggplot(aes(x = n, y = word, fill = title)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(vars(title), scales = "free_y") +
  labs(
    title = "Most frequent non-stopwords per Gutenberg book",
    x = "Count",
    y = NULL
  )
```

What happens:

-   We plot only the top 12 words for readability.
-   `facet_wrap()` produces one panel per book.
-   Students can compare lexical profiles side by side.

## Compare distinctive words with tf-idf {.scrollable}

**TF–IDF** (term frequency–inverse document frequency) scores how *characteristic* a word is for one text *relative to the whole set* you are comparing.

-   **Term frequency (TF)** — how often the word appears **in that book** (here reflected in the count column `n` after `count(title, word)`). A word that appears many times in *Alice* has high TF for *Alice*.
-   **Inverse document frequency (IDF)** — how *unusual* the word is **across books**. If a word appears in almost every title with similar counts, IDF is low (it is not distinguishing). If a word is concentrated in one book, IDF is higher for that contrast.
-   **TF × IDF** — the product (stored as `tf_idf` in `tidytext`) is high when the word is **both** frequent in that book **and** not equally typical everywhere else. That is why high `tf_idf` words are often proper names, place words, or topic-specific vocabulary, while raw “most frequent” lists are often dominated by shared literary English.

```{r}
#| label: gutenberg-tfidf

book_tfidf <- tokens |>
  count(title, word, sort = TRUE) |>
  bind_tf_idf(word, title, n) |>
  arrange(desc(tf_idf))

book_tfidf |>
  group_by(title) |>
  slice_head(n = 10) |>
  ungroup() |>
  select(title, word, n, tf_idf)
```

What happens:

-   `count(title, word)` builds a table of how many times each `word` appears in each `title` (document).
-   `bind_tf_idf(word, title, n)` adds three columns: `tf` (term frequency in that document), `idf` (inverse document frequency in the corpus), and `tf_idf` (their product).
-   We sort by `tf_idf` and show the top terms per book so you can read off **distinctive** words, not just frequent ones.
-   Interpretation: compare a high-`tf_idf` word’s role in the story (character, place, motif) with its `n` — a word can be distinctive with a moderate `n` if it barely appears in the other books.

## Analyze word pairs with bigrams (Hermann Hesse) {.scrollable}

Following the tidytext n-grams workflow, we now move from single words to **bigrams** (two consecutive words).\
Here we use three works by the German author **Hermann Hesse** from Project Gutenberg.

```{r}
#| label: gutenberg-bigrams-hesse

hesse_works <- gutenberg_works(author == "Hesse, Hermann") |>
  arrange(gutenberg_id) |>
  slice_head(n = 3)

hesse_books <- gutenberg_download(
  hesse_works$gutenberg_id,
  meta_fields = c("author", "title")
  )

hesse_bigrams <- hesse_books |>
  unnest_tokens(bigram, text, token = "ngrams", n = 2) |>
  separate_wider_delim(
    bigram,
    delim = " ",
    names = c("word1", "word2"),
    too_few = "align_start"
  ) |>
  filter(!is.na(word1), !is.na(word2)) |>
  anti_join(stop_words, by = join_by(word1 == word)) |>
  anti_join(stop_words, by = join_by(word2 == word))

hesse_bigram_counts <- hesse_bigrams |>
  count(title, word1, word2, sort = TRUE)

hesse_bigram_counts |>
  mutate(bigram = str_c(word1, word2, sep = " ")) |>
  group_by(title) |>
  slice_head(n = 10) |>
  ungroup() |>
  select(title, bigram, n)
```

What happens:

-   `unnest_tokens(..., token = "ngrams", n = 2)` creates overlapping two-word sequences.
-   `separate_wider_delim()` splits each bigram into `word1` and `word2`.
-   We remove stop words in both positions to focus on more informative pairs.
-   `count(title, word1, word2, sort = TRUE)` gives the most frequent word-pair patterns per text.

### Note on lexical diversity

In Exercise 5, lexical diversity is defined as:

`n_distinct(word) / n()`

-   `n_distinct(word)` = number of unique word forms (the vocabulary size in your token table).
-   `n()` = total number of tokens (all running words).
-   The ratio is often called **type-token ratio (TTR)**.

Interpretation:

-   Higher values mean a larger share of unique words (more varied vocabulary in that sample).
-   Lower values mean more repetition of the same words.
-   Important caveat: this measure is sensitive to text length (shorter texts often look more diverse), so compare texts of similar size or interpret with care.

## Exercises (Gutenberg + tidytext)

1.  Download one random Gutenberg text (choose one ID yourself), tokenize it, and compute the top 20 non-stopwords.
2.  Download from the Gutenberg archive the first 3 works by Wells, H. G. (Herbert George), create a token table for this mini-corpus, then build a three-book dataset (`bind_rows()`) and compare top words across books with a faceted bar plot.
3.  Compute tf-idf for the three-book dataset and interpret two high tf-idf words per book.
4.  Compute bigrams (`token = "ngrams", n = 2`) for one selected book and list the 20 most frequent bigrams after removing rows that contain stop words.
5.  For each of the three books, compute lexical diversity as `n_distinct(word) / n()` and compare the values.

# Part 3 — Working with XML data

We are working here with the data of the entire Lëtzebuerger Online Dictionary (LOD). The XML file of the entire dictionary can be downloaded from the Luxembourg open data portal ([LOD resource](https://data.public.lu/fr/datasets/letzebuerger-online-dictionnaire-lod-linguistesch-daten/#/resources/7aad8691-006f-43b0-8d65-e47e8d51696b)).

Conceptually, the XML workflow has four steps:

1.  Read the XML tree (`read_xml()`).
2.  Select `<entry>` nodes with XPath.
3.  Extract a few fields from each entry.
4.  Combine extracted rows into one tidy tibble.

## Read XML and inspect structure {.scrollable}

This block parses the XML file and checks basic structure.

```{r}
#| label: lod-read
#| message: false

lod_xml_path <- resolve_path("new_lod-art.xml")
lod_xml <- read_xml(lod_xml_path)

xml_name(lod_xml)
xml_find_all(lod_xml, "//entry") |> 
  length()
```

What happens:

-   `read_xml()` loads the XML tree into memory.
-   `xml_name()` identifies the root node (here: `lod`).
-   `xml_find_all(lod_xml, "//entry")` returns all dictionary entries anywhere in the tree.
-   `length()` gives the number of entries you could potentially process.

XPath quick reminder:

| XPath                                       | Meaning                      |
|-------------------------------------------|-----------------------------|
| `//entry`                                   | all entries                  |
| `.//lemma`                                  | lemma under current entry    |
| `.//ipa`                                    | IPA node under current entry |
| `.//targetLanguage[@lang='en']/translation` | English translations         |

Two key XML functions used below:

-   `xml_find_first(node, "xpath")` returns the **first matching XML node** (not plain text).
-   `xml_text(node)` returns the **text content inside that XML node**.

So this pattern:

`xml_find_first(entry_node, ".//lemma") |> xml_text()`

means: first find the `<lemma>` node inside one entry, then extract only its text value.

## Build a tidy tibble from XML {.scrollable}

This block converts XML nodes into a tidy one-row-per-entry table.

```{r}
#| label: lod-extract

# Full extraction over all entries is computationally heavy.
# We keep classroom render fast by using a sample.
entry_nodes <- xml_find_all(lod_xml, "//entry")
sample_n <- 1000
entry_idx <- sample.int(length(entry_nodes), size = sample_n)
entry_sample <- entry_nodes[entry_idx]

extract_entry <- function(entry_node) {
  # Pull all English translation strings for one entry.
  en_translations <- xml_find_all(
    entry_node,
    ".//meaning/targetLanguage[@lang='en']/translation"
  ) |>
    xml_text()

  # Return one-row tibble for this entry.
  tibble(
    lemma = xml_find_first(entry_node, ".//lemma") |> xml_text(),
    ipa = xml_find_first(entry_node, ".//ipa") |> xml_text(),
    n_en_translations = length(en_translations),
    first_en_translation = en_translations |> first(default = NA_character_)
  )
}

# takes some time
lod_entries <- entry_sample |>
  map(extract_entry) |>
  list_rbind()

lod_entries |> slice_head(n = 20)
```

What happens:

-   `entry_nodes` is an XML node set with all `<entry>` elements.
-   We randomly sample 1,000 entries so rendering stays fast.
-   `extract_entry()` runs on one XML node and returns one tibble row.
-   `xml_find_first()` gives an XML node object; `xml_text()` converts that node into a character string.
-   `map(extract_entry)` creates a list of small tibbles; `list_rbind()` stacks them into one dataset.
-   `first(..., default = NA_character_)` avoids an error when an entry has no English translation.

This block lists the most common first English translations.

```{r}
#| label: lod-top-en

lod_entries |>
  filter(!is.na(first_en_translation)) |>
  count(first_en_translation, sort = TRUE) |>
  slice_head(n = 20)
```

What happens:

-   Missing translations are filtered out.
-   Frequencies are counted and ranked.
-   This provides a quick lexical overview of common gloss labels in the sample.

## Exercises (XML / LOD)

1.  Extract a tibble with columns `lemma`, `ipa`, and `n_examples` (number of `<example>` nodes per entry).
2.  Create a table of the 25 most common first English translations and discuss why this summary may be linguistically limited.
3.  Sample 30 entries with at least one English translation and inspect whether `lemma` and `first_en_translation` appear semantically aligned.
4.  Create a summary table of average, median, and maximum `n_examples` across sampled entries.
5.  Build a filtered table of entries with both an IPA value and at least one English translation, and report the share of sampled entries that satisfy both conditions.
6.  Extract `partofspeech` for each sampled entry (together with `lemma`) and list the 20 most frequent part-of-speech labels.

# Solutions

Explained solutions for all exercise blocks are in `sessions/session-02-solutions.qmd`.
