---
title: "Session 2 — Data import and exploratory analysis"
subtitle: "Language Analytics in R"
author: "Peter Gilles"
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    code-fold: false
editor: visual
---

## Learning objectives {.scrollable}

By the end of this session you will:

-   Import tabular, textual, and XML language data into R.
-   Use tidyverse workflows for exploratory statistics and visualization.
-   Tokenize and analyze Gutenberg texts with `tidytext`.
-   Extract structured information from LOD XML with `xml2`.
-   Solve format-specific exercises for CSV, text, and XML data.

This session covers three common data formats for language analytics:

1.  **CSV** (tabular data)
2.  **Text** from Project Gutenberg (`tidytext` workflow)
3.  **XML** from the LOD dictionary

Each section ends with dedicated exercises. Explained solutions are provided in `sessions/session-02-solutions.qmd`.

# Setup

This block loads all packages used in this session and defines a helper function for robust file paths.

```{r}
#| label: setup
#| message: false

library(tidyverse)
library(tidytext)
library(gutenbergr)
library(xml2)

resolve_path <- function(filename) {
  if (file.exists(filename)) {
    filename
  } else {
    file.path("sessions", filename)
  }
}
```

What happens:

-   `library(...)` attaches the packages.
-   `resolve_path()` first checks local folder, then falls back to `sessions/`.
-   This avoids path errors when students render from different working directories.

# Part 1 — CSV data

## Read and inspect CSV {.scrollable}

We start with demographic data by commune in `rnrpp-population-commune.csv`.

This block reads the file into a tibble named `pop`.

```{r}
#| label: csv-read
#| message: false

csv_path <- resolve_path("rnrpp-population-commune.csv")
pop <- read_csv(
  csv_path,
  locale = locale(encoding = "ISO-8859-1"),
  show_col_types = FALSE
)

pop
```

What happens:

-   `resolve_path()` finds the file.
-   `read_csv(..., locale = locale(encoding = \"ISO-8859-1\"))` imports text with the correct source encoding.
-   Printing `pop` gives a quick first look at rows and columns.

This block performs structural inspection (shape and variable names).

```{r}
#| label: csv-inspect

glimpse(pop)
nrow(pop)
ncol(pop)
names(pop)
```

What happens:

-   `glimpse()` shows column types and sample values.
-   `nrow()` and `ncol()` report dataset size.
-   `names()` lists column names for later use in `mutate()` and `summarise()`.

## Basic exploratory statistics {.scrollable}

Create derived variables and inspect the largest communes.

```{r}
#| label: csv-explore

pop_totals <- pop |>
  mutate(
    total_population = FEMMES_MINEURES + HOMMES_MINEURS + FEMMES_MAJEURES + HOMMES_MAJEURS,
    adult_share = (FEMMES_MAJEURES + HOMMES_MAJEURS) / total_population
  )

pop_totals |>
  arrange(desc(total_population)) |>
  select(COMMUNE_NOM, total_population, adult_share) |>
  slice_head(n = 10)
```

What happens:

-   `mutate()` adds two analysis variables: total size and adult proportion.
-   `arrange(desc(...))` ranks communes from largest to smallest.
-   `select()` keeps only the most relevant columns for interpretation.

This block computes one-row summary statistics across all communes.

```{r}
#| label: csv-summary

pop_totals |>
  summarise(
    mean_commune_population = mean(total_population),
    median_commune_population = median(total_population),
    min_commune_population = min(total_population),
    max_commune_population = max(total_population)
  )
```

What happens:

-   `summarise()` reduces many rows to one row.
-   We compute central tendency (`mean`, `median`) and range (`min`, `max`).
-   This gives a fast statistical profile before plotting.

This block visualizes the top 20 communes with a horizontal bar chart.

```{r}
#| label: csv-plot
#| fig-width: 8
#| fig-height: 5

pop_totals |>
  arrange(desc(total_population)) |>
  slice_head(n = 20) |>
  mutate(COMMUNE_NOM = fct_reorder(COMMUNE_NOM, total_population)) |>
  ggplot(aes(x = total_population, y = COMMUNE_NOM)) +
  geom_col(fill = "steelblue") +
  labs(
    title = "Top 20 communes by population",
    x = "Population",
    y = NULL
  )
```

What happens:

-   `slice_head(n = 20)` keeps only the top communes.
-   `fct_reorder()` orders labels by population so the chart is readable.
-   `geom_col()` displays already-aggregated values.

## Exercises (CSV)

1.  Compute the total number of minors (`FEMMES_MINEURES + HOMMES_MINEURS`) per commune and list the top 15 communes.
2.  Remove the variable `COMMUNE_CODE` from the tibble.
3.  Add a variable `female_share` = total females / total population, then report the 10 communes with the highest `female_share`.
4.  Create a histogram of `total_population` and briefly describe its shape.
5.  Export the tibble `pop_totals` to a csv file using `write.csv`.

# Part 2 — Gutenberg text with tidytext

We use [`tidytext`](https://www.tidytextmining.com/tidytext) ideas: convert text to **one-token-per-row**, remove stop words, and compare frequencies.

## Download and tokenize Gutenberg texts {.scrollable}

This block downloads two books from Project Gutenberg, including title metadata.

```{r}
#| label: gutenberg-download

gutenberg_books <- gutenberg_download(
  c(11, 1342), # Alice in Wonderland, Pride and Prejudice
  meta_fields = c("author", "title")
 ) |>
  # bug in gutenberg; deduplication of rows necessary
  filter(
    is.na(lag(text)) | text != lag(text),
    .by = gutenberg_id
)

gutenberg_books
```

What happens:

-   `gutenberg_download()` retrieves cleaned text (header/footer removed).
-   Some Gutenberg texts (e.g., Alice on certain mirrors) include duplicated adjacent lines.
-   `filter(is.na(lag(text)) | text != lag(text), .by = gutenberg_id)` removes only consecutive duplicate lines within each book.
-   Each row is one line of text, with `gutenberg_id` and `title`.
-   The table is now ready for tokenization.

This block converts lines into words and removes stop words.

```{r}
#| label: gutenberg-tokenize

tokens <- gutenberg_books |>
  unnest_tokens(word, text) |>
  anti_join(stop_words, by = join_by(word))

tokens
```

What happens:

-   `unnest_tokens(word, text)` turns each line into multiple word rows.
-   Tokens are lowercase by default and punctuation is removed.
-   `anti_join()` removes common function words (`the`, `and`, etc.).

## Explore word frequencies {.scrollable}

This block counts how often each word appears in each title.

```{r}
#| label: gutenberg-count

top_words <- tokens |>
  count(title, word, sort = TRUE)

top_words |>
  group_by(title) |>
  slice_head(n = 15) |>
  ungroup()
```

What happens:

-   `count(title, word)` computes frequencies per book.
-   `sort = TRUE` returns highest counts first.
-   Grouping + `slice_head()` keeps the top words for each book.

This block visualizes top words by title in faceted panels.

```{r}
#| label: gutenberg-plot
#| fig-width: 9
#| fig-height: 5

top_words |>
  group_by(title) |>
  slice_head(n = 12) |>
  ungroup() |>
  mutate(word = fct_reorder(word, n)) |>
  ggplot(aes(x = n, y = word, fill = title)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(vars(title), scales = "free_y") +
  labs(
    title = "Most frequent non-stopwords per Gutenberg book",
    x = "Count",
    y = NULL
  )
```

What happens:

-   We plot only the top 12 words for readability.
-   `facet_wrap()` produces one panel per book.
-   Students can compare lexical profiles side by side.

## Compare distinctive words with tf-idf {.scrollable}

**TF–IDF** (term frequency–inverse document frequency) scores how *characteristic* a word is for one text *relative to the whole set* you are comparing.

-   **Term frequency (TF)** — how often the word appears **in that book** (here reflected in the count column `n` after `count(title, word)`). A word that appears many times in *Alice* has high TF for *Alice*.
-   **Inverse document frequency (IDF)** — how *unusual* the word is **across books**. If a word appears in almost every title with similar counts, IDF is low (it is not distinguishing). If a word is concentrated in one book, IDF is higher for that contrast.
-   **TF × IDF** — the product (stored as `tf_idf` in `tidytext`) is high when the word is **both** frequent in that book **and** not equally typical everywhere else. That is why high `tf_idf` words are often proper names, place words, or topic-specific vocabulary, while raw “most frequent” lists are often dominated by shared literary English.

```{r}
#| label: gutenberg-tfidf

book_tfidf <- tokens |>
  count(title, word, sort = TRUE) |>
  bind_tf_idf(word, title, n) |>
  arrange(desc(tf_idf))

book_tfidf |>
  group_by(title) |>
  slice_head(n = 10) |>
  ungroup() |>
  select(title, word, n, tf_idf)
```

What happens:

-   `count(title, word)` builds a table of how many times each `word` appears in each `title` (document).
-   `bind_tf_idf(word, title, n)` adds three columns: `tf` (term frequency in that document), `idf` (inverse document frequency in the corpus), and `tf_idf` (their product).
-   We sort by `tf_idf` and show the top terms per book so you can read off **distinctive** words, not just frequent ones.
-   Interpretation: compare a high-`tf_idf` word’s role in the story (character, place, motif) with its `n` — a word can be distinctive with a moderate `n` if it barely appears in the other books.

## Exercises (Gutenberg + tidytext)

1.  Download one additional Gutenberg text (choose one ID yourself), tokenize it, and compute the top 20 non-stopwords.
2.  Download from the Gutenberg archive all works by Jane Austen (check the help page for package `gutenbergr`.
3.  Build a three-book dataset (`bind_rows()`), then compare top words across books with a faceted bar plot.
4.  Compute tf-idf for the three-book dataset and interpret two high tf-idf words per book.
5.  Compute bigrams (`token = "ngrams", n = 2`) for one selected book and list the 20 most frequent bigrams after removing rows that contain stop words.
6.  For each of the three books, compute lexical diversity as `n_distinct(word) / n()` and compare the values.

# Part 3 — XML data (LOD dictionary)

The LOD XML file can be downloaded from the Luxembourg open data portal ([LOD resource](https://data.public.lu/fr/datasets/letzebuerger-online-dictionnaire-lod-linguistesch-daten/#/resources/7aad8691-006f-43b0-8d65-e47e8d51696b)).

Conceptually, the XML workflow has four steps:

1.  Read the XML tree (`read_xml()`).
2.  Select `<entry>` nodes with XPath.
3.  Extract a few fields from each entry.
4.  Combine extracted rows into one tidy tibble.

## Read XML and inspect structure {.scrollable}

This block parses the XML file and checks basic structure.

```{r}
#| label: lod-read
#| message: false

lod_xml_path <- resolve_path("new_lod-art.xml")
lod_xml <- read_xml(lod_xml_path)

xml_name(lod_xml)
xml_find_all(lod_xml, "//entry") |> 
  length()
```

What happens:

-   `read_xml()` loads the XML tree into memory.
-   `xml_name()` identifies the root node (here: `lod`).
-   `xml_find_all(lod_xml, "//entry")` returns all dictionary entries anywhere in the tree.
-   `length()` gives the number of entries you could potentially process.

XPath quick reminder:

| XPath                                       | Meaning                      |
|---------------------------------------------|------------------------------|
| `//entry`                                   | all entries                  |
| `.//lemma`                                  | lemma under current entry    |
| `.//ipa`                                    | IPA node under current entry |
| `.//targetLanguage[@lang='en']/translation` | English translations         |

Two key XML functions used below:

-   `xml_find_first(node, "xpath")` returns the **first matching XML node** (not plain text).
-   `xml_text(node)` returns the **text content inside that XML node**.

So this pattern:

`xml_find_first(entry_node, ".//lemma") |> xml_text()`

means: first find the `<lemma>` node inside one entry, then extract only its text value.

## Build a tidy tibble from XML {.scrollable}

This block converts XML nodes into a tidy one-row-per-entry table.

```{r}
#| label: lod-extract

# Full extraction over all entries is computationally heavy.
# We keep classroom render fast by using a sample.
entry_nodes <- xml_find_all(lod_xml, "//entry")
sample_n <- 1000
entry_idx <- sample.int(length(entry_nodes), size = sample_n)
entry_sample <- entry_nodes[entry_idx]

extract_entry <- function(entry_node) {
  # Pull all English translation strings for one entry.
  en_translations <- xml_find_all(
    entry_node,
    ".//meaning/targetLanguage[@lang='en']/translation"
  ) |>
    xml_text()

  # Return one-row tibble for this entry.
  tibble(
    lemma = xml_find_first(entry_node, ".//lemma") |> xml_text(),
    ipa = xml_find_first(entry_node, ".//ipa") |> xml_text(),
    n_en_translations = length(en_translations),
    first_en_translation = en_translations |> first(default = NA_character_)
  )
}

# takes some time
lod_entries <- entry_sample |>
  map(extract_entry) |>
  list_rbind()

lod_entries |> slice_head(n = 20)
```

What happens:

-   `entry_nodes` is an XML node set with all `<entry>` elements.
-   We randomly sample 1,000 entries so rendering stays fast.
-   `extract_entry()` runs on one XML node and returns one tibble row.
-   `xml_find_first()` gives an XML node object; `xml_text()` converts that node into a character string.
-   `map(extract_entry)` creates a list of small tibbles; `list_rbind()` stacks them into one dataset.
-   `first(..., default = NA_character_)` avoids an error when an entry has no English translation.

This block lists the most common first English translations.

```{r}
#| label: lod-top-en

lod_entries |>
  filter(!is.na(first_en_translation)) |>
  count(first_en_translation, sort = TRUE) |>
  slice_head(n = 20)
```

What happens:

-   Missing translations are filtered out.
-   Frequencies are counted and ranked.
-   This provides a quick lexical overview of common gloss labels in the sample.

## Exercises (XML / LOD)

1.  Extract a tibble with columns `lemma`, `ipa`, and `n_examples` (number of `<example>` nodes per entry).
2.  Create a table of the 25 most common first English translations and discuss why this summary may be linguistically limited.
3.  Sample 30 entries with at least one English translation and inspect whether `lemma` and `first_en_translation` appear semantically aligned.
4.  Create a summary table of average, median, and maximum `n_examples` across sampled entries.
5.  Build a filtered table of entries with both an IPA value and at least one English translation, and report the share of sampled entries that satisfy both conditions.
6.  Extract `partofspeech` for each sampled entry (together with `lemma`) and list the 20 most frequent part-of-speech labels.

# Solutions

Explained solutions for all exercise blocks are in `sessions/session-02-solutions.qmd`.
