---
title: "Session 1 — Introduction: R workflow and first steps"
subtitle: "Language Analytics in R"
author: "Peter Gilles"
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
editor: visual
---

## Learning objectives

By the end of this session you will:

-   Use RStudio and run R code in the console and in a Quarto document.
-   Create and inspect objects with the assignment operator (`<-`).
-   Write clear names and comments, and apply basic code style.
-   Call functions with correct syntax and use the pipe (`|>`).
-   Run and render a minimal Quarto (.qmd) script.
-   **Load and inspect text/language data** (e.g. CSV) and understand basics of data import for corpora.
-   **Use tidytext and quanteda** to turn short example text into tokens and a corpus, and run minimal counts.

This course focuses on **linguistic and language data in R**. Examples and exercises use words, frequencies, and simple text data so you can quickly apply the same ideas to your own corpora and experiments.

The material draws on [R for Data Science (2e)](https://r4ds.hadley.nz/) (chapters 2–4), [R for Linguists](https://eleanorchodroff.com/r-for-linguists/), and other linguistics-oriented R resources (see [Further reading](#further-reading)).

# RStudio and running R code

## The RStudio interface

-   **Console**: type R code and see results.
-   **Editor**: write and save scripts (e.g. `.qmd` or `.R`).
-   **Environment**: view objects you have created (datasets, variables).
-   **Files / Plots / Help**: navigate files, view plots, read help.

Run code from the editor with **Cmd+Enter** (Mac) or **Ctrl+Enter** (Windows/Linux).

## Coding basics {#sec-basics}

You can use R as a calculator (e.g. for reaction times, accuracy, or counts):

```{r}
#| label: calc-basics

1 / 200 * 30
(59 + 73 + 2) / 3
sin(pi / 2)
```

Create objects with **assignment**:

```{r}
#| label: assignment

# e.g. store a value like VOT (ms) or word length
vot_ms <- 25
word_length <- nchar("linguistics")
```

The value is stored but not printed. Type the object name to inspect it:

```{r}
vot_ms
word_length
```

Combine values into a **vector** with `c()` — like a column of observations (e.g. reaction times, word lengths, or a list of words):

```{r}
# numeric vector: e.g. reaction times in ms
rt_ms <- c(320, 415, 380, 290, 410)

# character vector: e.g. word forms or conditions
words <- c("cat", "dog", "mouse")
```

Every assignment has the form:

``` text
object_name <- value
```

Read it as: *“object name gets value”*.

**Tip:** In RStudio use **Alt + -** (minus) to insert `<-` with spaces.

# Comments and names

## Comments

R ignores everything after `#` on a line. Use comments to explain *why* you did something (e.g. why you chose a threshold or excluded a condition), not just *what* the code does.

```{r}
#| label: comments

# word list for lexical decision practice
words <- c("cat", "dog", "mouse")

# mean reaction time (ms)
mean(rt_ms)
```

## What’s in a name? {#sec-names}

-   Names must **start with a letter** and contain only letters, numbers, `_`, and `.`.
-   Use **snake_case**: lowercase words separated by `_` (e.g. `word_list`, `response_time`).

```{r}
#| label: naming

response_time_ms <- 320
word_list <- c("the", "cat", "sat")
# Avoid: responseTimeMs, Word.List
```

R is case-sensitive and does not fix typos: `word_list` and `Word_list` are different objects.

**Tip:** Type a prefix and press **Tab** for autocomplete; use **↑** or **Cmd+↑** to recall previous commands.

# Calling functions

Functions are called like this:

``` text
function_name(argument1 = value1, argument2 = value2, ...)
```

Example: `seq()` creates sequences (e.g. for item IDs or conditions):

```{r}
#| label: seq-call

seq(from = 1, to = 10)
seq(1, 5, by = 0.5)
```

You can omit argument names for the first arguments. Use **Tab** after typing a function name to see arguments and help.

**Strings** (text) need **matched quotation marks** — essential when working with words and corpora:

```{r}
word <- "hello"
# Compare strings (e.g. for accuracy coding)
"linguistics" == "linguistics"
"linguistics" == "lingiustics"
```

If you see a `+` in the console, R is waiting for more input (often a missing `"` or `)`). Press **Escape** to cancel.

# Code style {#sec-style}

Good style makes code easier to read for you and others. We follow the [tidyverse style guide](https://style.tidyverse.org/).

## Names (reminder)

-   Use lowercase, numbers, and `_`.
-   Prefer **long, descriptive names** (e.g. `participant_response_time` over `prt`).

## Spaces

-   Put spaces around `+`, `-`, `==`, `<`, etc. (except `^`).
-   Put spaces around `<-`.
-   Put a space after every comma.

```{r}
#| label: style-spaces

# Strive for: spaces around operators and after comma
n_char <- nchar("word")
mean(rt_ms, na.rm = TRUE)
# Avoid: n_char<-nchar("word") or mean(rt_ms,na.rm=TRUE)
```

## Pipes

The pipe `|>` passes the result on the left into the function on the right. Read it as *“then”*.

-   Put a space before `|>` and put `|>` at the **end** of the line.
-   Indent the next line by two spaces.

We’ll use the pipe with **language data** in the next section.

## Sectioning comments

Break long scripts into sections:

``` r
# Load data --------------------------------------

# Plot data --------------------------------------
```

In RStudio: **Cmd+Shift+R** (Mac) or **Ctrl+Shift+R** (Windows/Linux) inserts a section header.

**Tip:** Install the **styler** package and use the command palette (**Cmd+Shift+P**) to format code automatically.

# Working with text data and corpora (data import)

For language analytics we usually work with **tables** (e.g. CSV or Excel) or **text files**. Here we cover reading tabular data with **readr**, then **tidytext** and **quanteda** for turning text into a format you can analyze. Later sessions will go deeper (frequency lists, sentiment, document classification, etc.).

## Packages

Install a package once with `install.packages("package_name")`. Load it every time you start R or open a document with `library(package_name)`:

```{r}
#| label: packages
#| message: false

# Core packages for data import and manipulation
library(tidyverse)
```

## Reading a CSV file

Use `read_csv()` from the **readr** package (included in tidyverse) to read comma-separated files — e.g. experiment results, word lists, or corpus metadata:

```{r}
#| label: read-csv

# Example: create a small toy CSV in memory (simulating a saved file)
toy_csv <- "word,freq,pos
the,100,det
cat,5,noun
sat,3,verb
the,100,det"

# read_csv() can take a string or a file path
word_freq <- read_csv(toy_csv, show_col_types = FALSE)
word_freq
```

In practice you will use a **file path** to your data, e.g. `read_csv("data/lexical_decision.csv")`. Put your data files in a project folder (e.g. `data/`) and use relative paths so your project stays portable.

## Inspecting the dataset

-   **Print** the data frame: type its name.
-   **View()**: opens a spreadsheet-like viewer in RStudio.
-   **glimpse()**: compact view of columns and types (very useful for many variables).

```{r}
#| label: inspect-data

glimpse(word_freq)
nrow(word_freq)
ncol(word_freq)
names(word_freq)
```

## Accessing columns

Use `$` to get one column as a vector (e.g. for computing over word frequencies or reaction times):

```{r}
word_freq$word
word_freq$freq
mean(word_freq$freq)
```

## Reading text with tidytext {#reading-text-with-tidytext}

The **tidytext** package treats text as a **tidy** data frame: one row per token (e.g. word). You start from a table with one row per document (or line) and a text column, then **tokenize** with `unnest_tokens()`.

**Install once:** `install.packages("tidytext")`. Then load it with `library(tidytext)` (and `library(tidyverse)` for tibbles and the pipe).

**Example:** build a small “corpus” as a tibble with document IDs and text, then split into words:

```{r}
#| label: tidytext-example
#| message: false

library(tidytext)

# Toy corpus: one row per document
toy_corpus_tidy <- tibble(
  doc_id = c("doc1", "doc2", "doc3"),
  text = c(
    "The cat sat on the mat.",
    "The dog ran in the sun.",
    "A cat and a dog."
  )
)

# Tokenize: one row per word (punctuation removed, lowercased by default)
toy_tokens <- toy_corpus_tidy |>
  unnest_tokens(output = word, input = text)

toy_tokens
```

You can then count words, filter stopwords, or join with lexicons (e.g. sentiment). Word counts:

```{r}
toy_tokens |>
  count(word, sort = TRUE)
```

For real projects you might read documents with `read_csv()` or `read_lines()` into a tibble with a `text` column, then use `unnest_tokens()` the same way.

## Reading text with quanteda {#reading-text-with-quanteda}

The **quanteda** package works with a **corpus** object: a collection of documents with metadata. You can build a corpus from a character vector (one string per document) or from a data frame with a text column.

**Install once:** `install.packages("quanteda")`. Then load with `library(quanteda)`.

**Example:** use the same toy sentences; create a corpus and inspect it:

```{r}
#| label: quanteda-example
#| message: false

library(quanteda)

# Option A: corpus from a data frame (need docid and text column)
toy_df <- tibble(
  doc_id = c("doc1", "doc2", "doc3"),
  text = c(
    "The cat sat on the mat.",
    "The dog ran in the sun.",
    "A cat and a dog."
  )
)
corp <- corpus(toy_df, text_field = "text")

# Option B: corpus from a named character vector
# texts <- c(doc1 = "The cat sat on the mat.", doc2 = "The dog ran.", ...)
# corp <- corpus(texts)

# Basic inspection
summary(corp)
ndoc(corp)
ntoken(corp)
```

To get a **document-term matrix** (documents × terms) for frequency analysis or modelling:

```{r}
# Tokenize and build document-term matrix
dtm <- corp |>
  tokens(remove_punct = TRUE) |>
  dfm()
dtm
```

Later you can use `readtext::readtext()` to read files from disk into a data frame, then pass that to `corpus(..., text_field = "text")`.

**When to use which?**

-   **tidytext**: you prefer working in tibbles and dplyr (one row per token); good for word counts, sentiment, and tidy-style workflows.
-   **quanteda**: you need a corpus object, document-term matrices, or built-in NLP (e.g. keywords, collocations, classification); good for corpus linguistics and text mining.
-   **readr/read_csv**: for tabular data (responses, word lists, metadata); combine with tidytext or quanteda when your “text” lives in a column.

# First taste of data transformation (with language data) {#first-taste-of-data-transformation-with-language-data}

We use **dplyr** (part of the tidyverse) to filter, select, and summarize data. Key ideas:

-   First argument is always a **data frame**.
-   You refer to **column names** without quotes.
-   The result is a **new data frame** (the original is unchanged).

Use the pipe to chain steps. Example: take a small **word/frequency** table, filter by frequency, then sort:

```{r}
#| label: first-pipe
#| message: false

# Build a tiny “corpus-like” table: word, frequency, condition
word_data <- tibble(
  word = c("the", "cat", "sat", "on", "mat", "dog", "ran"),
  freq = c(100, 5, 3, 50, 2, 8, 4),
  pos = c("det", "noun", "verb", "prep", "noun", "noun", "verb")
)

# Then: filter, arrange, show
word_data |>
  filter(freq >= 4) |>
  arrange(desc(freq)) |>
  select(word, freq, pos)
```

Read this as: *“Take `word_data`, then keep rows where `freq` is at least 4, then sort by `freq` descending, then show columns `word`, `freq`, `pos`.”*

Main verbs you will see again in Session 2:

-   `filter()` — keep rows that satisfy a condition (e.g. certain words or participants).
-   `select()` — keep or drop columns.
-   `mutate()` — add or change columns (e.g. log frequency, word length).
-   `arrange()` — sort rows.
-   `summarize()` — compute summaries (often with `group_by()`).

# Scripting in Quarto

## Why Quarto?

-   Combine **text, code, and output** in one document — ideal for reporting on corpus or experiment results.
-   **Reproducible**: re-run and re-render to update tables and figures.
-   Output to HTML, PDF, or other formats.

## Running code in a .qmd file

-   **Code chunks** are enclosed in ```` ```{r} ```` and ```` ``` ````.
-   Run one chunk: **Cmd+Enter** (Mac) or **Ctrl+Enter** (Windows/Linux).
-   Render the whole document: **Cmd+Shift+K** (or **Ctrl+Shift+K**) or click *Render*.

Example chunk options:

``` r
#| label: my-chunk
#| echo: true
#| eval: true
```

-   `label`: unique name for the chunk.
-   `echo`: whether to show the code in the output.
-   `eval`: whether to run the code.

# Hands-on exercises (toy language data)

1.  **Typo.** Why does this fail? (Look very carefully at the object name.)

    ``` r
    my_variable <- 10
    my_varıable   # error (hint: different character than 'i')
    ```

2.  **Vectors and strings.** In a new code chunk:

    -   Create a character vector `test_words` with at least three words (e.g. `c("word1", "word2", "word3")`).
    -   Use `length(test_words)` and `nchar(test_words)` and interpret the result.

3.  **Comparisons.** Run these and say what they return (TRUE/FALSE) and why:

    ``` r
    "dog" == "dog"
    "dog" != "cat"
    nchar("linguistics") > 5
    ```

4.  **Data import and inspection.** Create a short CSV string with columns `word` and `count` (e.g. three rows). Read it with `read_csv()`, then use `glimpse()` and `$` to inspect the `count` column and compute `mean(count)`.

5.  **Pipe and filter.** Using the `word_data` table from the [First taste](#first-taste-of-data-transformation-with-language-data) section (re-run that chunk if needed), write a pipe that:

    -   keeps only rows where `pos` is `"noun"`,
    -   then arranges by `freq` ascending.

6.  **Style.** Restyle this into a clear, multi-line pipe (spaces, line breaks, indentation):

    ``` r
    word_data|>filter(freq>3)|>arrange(freq)|>select(word,freq)
    ```

7.  **tidytext and quanteda (hands-on).** Use the same toy texts as in the [tidytext](#reading-text-with-tidytext) and [quanteda](#reading-text-with-quanteda) sections (or two short sentences of your own in a tibble with columns `doc_id` and `text`).

    **(a) tidytext:** Tokenize the text with `unnest_tokens(output = word, input = text)`. Then, using the pipe, count words with `count(word, sort = TRUE)` and show the result. How many times does *the* appear?

    **(b) quanteda:** Build a corpus with `corpus(your_tibble, text_field = "text")`. Call `summary(corp)` and `ntoken(corp)`. How many tokens are in the corpus? Optional: build a document-term matrix with `tokens(corp, remove_punct = TRUE) |> dfm()` and print it.

    **(c)** In one sentence: what is the main difference between the tidytext output (tibble of tokens or counts) and the quanteda `dfm` (document-term matrix)?

# Summary

-   **Run R** in the console and in Quarto chunks; use **Cmd+Enter** and **Render**.
-   **Assign** with `<-`; use **snake_case** and **comments** for clarity.
-   **Call functions** with `name(arg = value)`; use **Tab** and Help.
-   **Style:** spaces, pipe at end of line, sectioning comments; use **styler** and **Cmd+Shift+R**.
-   **Quarto:** combine text and code; re-render for reproducible reports.
-   **Data import:** use `read_csv()` for CSV tables; use `glimpse()`, `names()`, and `$` to inspect; keep data in a `data/` folder and use relative paths. Use **tidytext** (tibble + `unnest_tokens()`) for one-token-per-row workflows and **quanteda** (`corpus()`, `tokens()`, `dfm()`) for corpus objects and document-term matrices.
-   **dplyr** and the pipe will be used throughout the course for linguistic and corpus data; we go deeper in Session 2.

# Further reading {#further-reading}

**R and workflow**

-   [R4DS — Workflow: basics (Ch 2)](https://r4ds.hadley.nz/workflow-basics.html)
-   [R4DS — Data transformation (Ch 3)](https://r4ds.hadley.nz/data-transform.html)
-   [R4DS — Workflow: code style (Ch 4)](https://r4ds.hadley.nz/workflow-style.html)
-   [tidyverse style guide](https://style.tidyverse.org/)
-   [Quarto documentation](https://quarto.org/docs/guide/)

**R for linguists and language data**

-   [R for Linguists](https://eleanorchodroff.com/r-for-linguists/) — basics, variables, vectors, datasets, dplyr, ggplot2, strings and tidyr (with linguistic examples and datasets).
-   [A&S500 / Lin517 — Quantitative Investigations](https://jofrhwld.github.io/AandS500_2023/) (J. Fruehwald) — Quarto, strings, text, sentiment, quanteda.
-   [Lin611 — Quantitative Methods in Linguistics](https://jofrhwld.github.io/2023_Lin611/) (J. Fruehwald) — tidyverse, Winter (2019) textbook, OSF data, vowel normalization, modelling.
-   [NTNU ENC2036 — Corpus Linguistics](https://alvinntnu.github.io/NTNU_ENC2036_LECTURES/) — corpus methods, R packages (tidytext, quanteda, stringr, etc.).
-   [LSA 2017 — Linguistic data in R](https://jofrhwld.github.io/teaching/courses/2017_lsa/) — further linguistics-oriented R materials.
-   [Applied statistics for linguists](https://appliedstatisticsforlinguists.org/) — Winter, statistics and R for linguists (including proofs/chapters).
