---
title: "Session 1 — Introduction: R workflow and first steps"
subtitle: "Language Analytics in R"
author: "Peter Gilles"
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    code-fold: false
  revealjs:
    output-file: session-01-slides.html
    slide-level: 2
    smaller: true
    incremental: true
    transition: slide
    scrollable: true
    include-in-header:
      - text: |
          <style>
          /* Scrollable slides: constrain height so overflow can scroll (see SO 76962902) */
          .reveal .slide.scrollable {
            height: 100% !important;
            max-height: 100vh !important;
            overflow-y: auto !important;
            overflow-x: hidden;
            box-sizing: border-box;
          }
          </style>
editor: visual
---

## Learning objectives {.scrollable}

By the end of this session you will:

-   Use RStudio and run R code in the console and in a Quarto document.
-   Create and inspect objects with the assignment operator (`<-`).
-   Write clear names and comments, and apply basic code style.
-   Call functions with correct syntax and use the pipe (`|>`).
-   Run and render a minimal Quarto (.qmd) script.
-   **Load and inspect text/language data** (e.g. CSV) and understand basics of data import for corpora.

This course focuses on **linguistic and language data in R**. Examples and exercises use words, frequencies, and simple text data so you can quickly apply the same ideas to your own corpora and experiments.

The material draws on [R for Data Science (2e)](https://r4ds.hadley.nz/) (chapters 2–4), [R for Linguists](https://eleanorchodroff.com/r-for-linguists/), and other linguistics-oriented R resources (see [Further reading](#further-reading)).

# RStudio and running R code

## The RStudio interface

-   **Console**: type R code and see results.
-   **Editor**: write and save scripts (e.g. `.qmd` or `.R`).
-   **Environment**: view objects you have created (datasets, variables).
-   **Files / Plots / Help**: navigate files, view plots, read help.

Run code from the editor with **Cmd+Enter** (Mac) or **Ctrl+Enter** (Windows/Linux).

## Getting help

Prefix a command with `?`, e.g. `?mean` to call the help page for this command. Or use the `Help` pane in RStudio.

```{r}
# Access help for a function by name:
?mean

# Alternatively, use the help() function:
help("mean")

# Search help topics with apropos() or help.search():
apropos("lin")  # Shows all objects/functions with "lin" in the name
help.search("linear model")  # Looks up help topics related to linear models

# View example usage for functions:
example(mean)

# See the vignette (long-form documentation) for a package:
# vignette("dplyr")
```

There are numerous R [cheatsheets](https://posit.co/resources/cheatsheets/) that summarise command usages in a structured way.

## Packages

Install a package once with `install.packages("package_name")` or use the `packages` tab in RStudio. Load it every time you start R or open a document with `library(package_name)`:

```{r}
#| label: packages
#| message: false

# Core packages for data import and manipulation
library(tidyverse)
```

## 

## Coding basics {#sec-basics}

You can use R as a calculator (e.g. for reaction times, accuracy, or counts):

```{r}
#| label: calc-basics

1 / 200 * 30
(59 + 73 + 2) / 3
sin(pi / 2)
```

Create objects with **assignment**:

```{r}
#| label: assignment

calc <- (59 + 73 + 2) / 3

# e.g. store a value like VOT (ms) or word length
vot_ms <- 25
word_length <- nchar("linguistics")
```

Every assignment has the form:

``` text
object_name <- value
```

Read it as: *“object name gets value”*.

**Tip:** In RStudio use **Alt + -** (minus) to insert `<-` with spaces.

The value is stored but not printed. Type the object name to inspect it:

```{r}
vot_ms
word_length
```

Combine values into a **vector** with `c()` — like a column of observations (e.g. reaction times, word lengths, or a list of words):

```{r}
#| label: vectors-c
# numeric vector: e.g. reaction times in ms
rt_ms <- c(320, 415, 380, 290, 410)

# character vector: e.g. word forms or conditions
words <- c("cat", "dog", "mouse")

# a rose is a rose — last expression is printed in the report
stein <- c("a", "rose", "is", "a", "rose", "is", "a", "rose")
print(stein)
```

Instead of referencing every token in a vector individually, we can convert them to **factor**s.

```{r}
stein.fac <- as.factor(c("a", "rose", "is", "a", "rose", "is", "a", "rose"))
print(stein.fac)
```

Print as table

```{r}
table(stein.fac)
```

# Tibbles and data frames {.scrollable}

Vectors and factors hold a single column of values. For **tabular data** (rows and columns), R uses **data frames**. A **tibble** is a modern, tidyverse-style data frame: it prints nicely, doesn’t convert strings to factors by default, and works well with the pipe and dplyr.

## Creating a tibble

Use **`tibble()`** to build a table from named vectors (each vector becomes a column). Columns can be referred to by name in later code.

```{r}
#| label: tibble-intro

library(tibble)

# Example: a small table (e.g. word, frequency, part of speech)
lex <- tibble(
  word = c("the", "cat", "sat", "on", "mat"),
  freq = c(100, 5, 3, 50, 2),
  pos  = c("det", "noun", "verb", "prep", "noun")
)
lex
```

With **`data.frame()`** you get a classic R data frame; tibbles are usually preferred in the tidyverse because of their print format and consistent behaviour.

## Inspecting structure

-   **Print:** typing the name shows the first rows and column types (tibbles don’t flood the console).
-   **`glimpse()`** (dplyr): one row per column, with type and a few values.
-   **`nrow()`, `ncol()`, `names()`**: number of rows, number of columns, column names.

```{r}
#| label: tibble-inspect

glimpse(lex)
nrow(lex)
names(lex)
```

## Accessing columns

Use **`$`** to get one column as a vector: `lex$word`, `lex$freq`. You can use that in expressions (e.g. `mean(lex$freq)`) or pass it to functions. Later you’ll use dplyr verbs (`filter`, `select`, `mutate`) to work with whole tables without pulling columns out by hand.

```{r}
lex$word
mean(lex$freq)
```

# Comments and names

## Comments {.scrollable}

R ignores everything after `#` on a line. Use comments to explain *why* you did something (e.g. why you chose a threshold or excluded a condition), not just *what* the code does.

```{r}
#| label: comments

# word list for lexical decision practice
words <- c("cat", "dog", "mouse")

# mean reaction time (ms)
mean(rt_ms)
```

## Naming conventions {#sec-names .scrollable}

-   Names must **start with a letter** and contain only letters, numbers, `_`, and `.`.
-   Use **snake_case**: lowercase words separated by `_` (e.g. `word_list`, `response_time`).

```{r}
#| label: naming

response_time_ms <- 320
word_list <- c("the", "cat", "sat")
# Avoid: responseTimeMs, Word.List
```

R is case-sensitive and does not fix typos: `word_list` and `Word_list` are different objects.

**Tip:** Type a prefix and press **Tab** for autocomplete; use **↑** or **Cmd+↑** to recall previous commands.

# Calling functions

Functions are called like this:

``` text
function_name(argument1 = value1, argument2 = value2, ...)
```

Example: `seq()` creates sequences (e.g. for item IDs or conditions):

```{r}
#| label: seq-call

seq(from = 1, to = 10)
seq(1, 5, by = 0.5)
```

You can omit argument names for the first arguments. Use **Tab** after typing a function name to see arguments and help.

**Strings** (text) need **matched quotation marks** — essential when working with words and corpora:

```{r}
# quotation marks around strings
word <- "hello"

# Compare strings with == (e.g. for accuracy coding)
"linguistics" == "linguistics"
"linguistics" == "lingiustics"
```

If you see a `+` in the console, R is waiting for more input (often a missing `"` or `)`). Press **Escape** to cancel.

# Code style {#sec-style}

Good style makes code easier to read for you and others. We follow the [tidyverse style guide](https://style.tidyverse.org/).

## Names (reminder) {.scrollable}

-   Use lowercase, numbers, and `_`.
-   Prefer **long, descriptive names** (e.g. `participant_response_time` over `prt`).

## Spaces {.scrollable}

-   Put spaces around `+`, `-`, `==`, `<`, etc. (except `^`).
-   Put spaces around `<-`.
-   Put a space after every comma.

```{r}
#| label: style-spaces

# Strive for: spaces around operators and after comma
n_char <- nchar("word")
mean(rt_ms, na.rm = TRUE)
# Avoid: n_char<-nchar("word") or mean(rt_ms,na.rm=TRUE)
```

## Pipes {.scrollable}

The pipe `|>` (alternatively `%>%` when using package `magrittr`) passes the result on the left into the function on the right. Read it as *“then”*.

-   Put a space before `|>` and put `|>` at the **end** of the line.
-   Indent the next line by two spaces.

```{r}
c("Méindeg", "Dënschdeg", "Mëttwoch") |>
  nchar()
```

We’ll use the pipe with **language data** in the next section.

## Sectioning comments {.scrollable}

Break long scripts into sections:

``` r
# Load data --------------------------------------

# Plot data --------------------------------------
```

In RStudio: **Cmd+Shift+R** (Mac) or **Ctrl+Shift+R** (Windows/Linux) inserts a section header.

**Tip:** Install the **styler** package and use the command palette (**Cmd+Shift+P**) to format code automatically.

# Scripting in Quarto

## Why Quarto? {.scrollable}

-   Combine **text, code, and output** in one document — ideal for reporting on corpus or experiment results.
-   **Reproducible**: re-run and re-render to update tables and figures.
-   Output to HTML, PDF, or other formats.

## Running code in a .qmd file {.scrollable}

-   Create an empty Quarto file in RStudio: File \> New File \> Quarto Document ...
-   **Code chunks** are enclosed in ```` ```{r} ```` and ```` ``` ````.
-   Run one chunk: **Cmd+Enter** (Mac) or **Ctrl+Enter** (Windows/Linux).
-   Render the whole document: **Cmd+Shift+K** (or **Ctrl+Shift+K**) or click *Render*.

Example chunk options:

``` r
#| label: my-chunk
#| echo: true
#| eval: true
```

-   `label`: unique name for the chunk.
-   `echo`: whether to show the code in the output.
-   `eval`: whether to run the code.

# Working with text data and corpora (data import)

For language analytics we usually work with **tables** (e.g. CSV or Excel) or **text files**. Here we focus on reading tabular data with **readr** and tidyverse tools. We will cover **tidytext** and **quanteda** in Session 3 (corpus analysis).

## Reading a CSV file {.scrollable}

Use `read_csv()` from the **readr** package (included in tidyverse) to read comma-separated files. Example: population by commune (Luxembourg), file `rnrpp-population-commune.csv` (same folder as this document, or adjust the path if needed).

```{r}
#| label: read-csv
#| echo: false

# Path: same folder as this .qmd, or "sessions/rnrpp-population-commune.csv" from project root
pop_path <- "rnrpp-population-commune.csv"
pop <- read_csv(pop_path, show_col_types = FALSE)
pop
```

## Inspecting the dataset {.scrollable}

-   **Print** the data frame: type its name.
-   **View(pop)**: opens a spreadsheet-like viewer in RStudio.
-   **glimpse()**: compact view of columns and types.

```{r}
#| label: inspect-data

glimpse(pop)
nrow(pop)
ncol(pop)
names(pop)
head(pop, 5)
```

## Accessing columns {.scrollable}

Use `$` to get one column as a vector (e.g. to compute totals or means):

```{r}
#| label: access-columns

pop$COMMUNE_NOM
pop$FEMMES_MAJEURES
# Mean population per commune (total = sum of the four count columns)
mean(pop$FEMMES_MINEURES + pop$HOMMES_MINEURS + pop$FEMMES_MAJEURES + pop$HOMMES_MAJEURS)
```

## Aggregating the data {.scrollable}

With **dplyr** you can **aggregate** by groups (e.g. totals, means). Example: add a total population per commune, then show the top 10 communes by total population; or summarize across all rows (e.g. national totals by category).

```{r}
#| label: aggregate-pop

# Total population per commune (using the pipe)
pop |>
  mutate(
    total = FEMMES_MINEURES + HOMMES_MINEURS + FEMMES_MAJEURES + HOMMES_MAJEURS
  ) |>
  arrange(desc(total)) |>
  select(COMMUNE_NOM, total) |>
  head(10)
```

```{r}
#| label: aggregate-summarize

# One-row summary: national totals by category (sum across all communes)
pop |>
  summarise(
    femmes_mineures = sum(FEMMES_MINEURES),
    hommes_mineurs  = sum(HOMMES_MINEURS),
    femmes_majeures = sum(FEMMES_MAJEURES),
    hommes_majeurs  = sum(HOMMES_MAJEURS),
    total          = sum(FEMMES_MINEURES + HOMMES_MINEURS + FEMMES_MAJEURES + HOMMES_MAJEURS)
  )
```

So: **load** with `read_csv()`, **inspect** with `glimpse()`, `names()`, `head()`, **access** columns with `$`, and **aggregate** with `mutate()`, `summarise()`, `arrange()`, and `select()` (and later `group_by()` for group-wise summaries).

For text mining workflows with **tidytext** and **quanteda**, see **Session 3**.

# Import LOD dictionary data (XML)

## Getting the file {.scrollable}

The LOD linguistic dataset is published as **open data** by the Luxembourg government (Zenter fir d'Lëtzebuerger Sprooch) under **CC0 (**[LOD – Linguistesch Daten (resource)](https://data.public.lu/fr/datasets/letzebuerger-online-dictionnaire-lod-linguistesch-daten/#/resources/7aad8691-006f-43b0-8d65-e47e8d51696b). Download the zip, extract it, and place the XML file (`new_lod-art.xml`, \~74 MB) in a folder in your project (e.g. `data/`). Below we read it from that folder and inspect its structure.

**Install once:** `install.packages("xml2")`. Then load with `library(xml2)`.

## Reading the XML file {.scrollable}

Read the XML from the file path. The file is large, so the first parse may take a few seconds.

```{r}
#| label: lod-read-xml
#| message: false

library(xml2)

# Path to the XML file (e.g. in data/; adjust if your file is elsewhere)
lod_xml_path <- "new_lod-art.xml"

# Read XML from the folder
lod_xml <- read_xml(lod_xml_path)

# Root element
xml_name(lod_xml)
xml_children(lod_xml) |> length()   # number of top-level children (entries)
```

## Structure of the LOD XML {.scrollable}

The document has a single root element `<lod>`, which contains many **`<entry>`** elements. Each entry represents one **headword** (lemma) and its dictionary information. The hierarchy is:

**1. Entry level (`<entry id="...">`)**

-   **`<lemma>`** — the headword (e.g. "A").
-   **`<ipa>`** — IPA pronunciation (e.g. "aː").
-   **`<microStructure>`** — grammatical and semantic information.

**2. Inside `<microStructure>`**

-   **`<partOfSpeech gen="...">`** — part of speech (e.g. SUBST, with optional gender).
-   **`<inflection type="...">`** — inflectional forms (e.g. plural), with `<form nRuleForm="...">` for each form.
-   **`<grammaticalUnit>`** — groups one or more **meanings**.

**3. Meaning level (`<meaning id="..." video="...">`)**

-   **`<number>`** — sense number.
-   **`<targetLanguage lang="de|fr|en|pt|nl">`** — translations:
    -   **`<translation>`** — equivalent in the target language.
    -   **`<semanticClarifier>`** — short clarification (e.g. “Sehorgan” for “Auge”).
-   **`<examples count="...">`** — example sentences:
    -   **`<example>`** (optional `id`):
        -   **`<text>`** — Luxembourgish example: `<word>` elements for each token, `<inflectedHeadword>` for the headword form used; optional `<attribute>` (e.g. EGS = example).
        -   **`<gloss>`** (optional) — gloss/paraphrase, again with `<word>` elements.

So each **entry** → **microStructure** → **grammaticalUnit** → one or more **meaning** → translations in several languages and **examples** (with optional glosses).

## How XPath works

**XPath** is a language for selecting nodes in an XML (or HTML) tree. In R you use it with **xml2** via `xml_find_first()` (one node) and `xml_find_all()` (all matching nodes). The expression is a path that describes where to go in the tree.

| XPath | Meaning | Example |
|------------------------|------------------------|------------------------|
| `nodename` | Direct child with that name | `lemma` → children of the current node named `lemma` |
| `/nodename` | Child from the **root** | `/lod` → root element; `/lod/entry` → every `entry` directly under `lod` |
| `//nodename` | **Any** descendant (any depth) | `//entry` → every `entry` in the document |
| `.//nodename` | Any descendant **from the current node** | Inside an entry, `.//lemma` finds `lemma` anywhere under that entry |
| `*` | Any element (wildcard) | `./*` → all direct children of the current node |
| `[@attr='value']` | Filter by attribute | `targetLanguage[@lang='de']` → `targetLanguage` whose `lang` is `de` |
| `path1/path2` | Then: path2 under path1 | `.//meaning/examples/example` → `example` under `examples` under `meaning` |

**Examples on the LOD:**

-   `//entry` — every entry in the document (from root).
-   `//entry/lemma` — every `lemma` that is a direct child of an `entry`.
-   `.//lemma` — the lemma of the **current** entry (when you’re already on one entry).
-   `.//meaning/targetLanguage[@lang='en']/translation` — English translations under the current entry.
-   `.//meaning/examples/example` — all example nodes under the current entry.

The **dot** (`.`) is the current node; **`.//`** means “starting from here, any descendant”. Without the dot, `//` starts from the document root.

## Inspecting one entry in R {.scrollable}

To see the structure of a single entry and extract a few fields:

```{r}
#| label: lod-structure

# First entry
first_entry <- xml_find_first(lod_xml, "//entry")

# Lemma and IPA
xml_find_first(first_entry, ".//lemma") |> xml_text()
xml_find_first(first_entry, ".//ipa")    |> xml_text()

# Part of speech
xml_find_first(first_entry, ".//partOfSpeech") |> xml_text()
xml_find_first(first_entry, ".//partOfSpeech") |> xml_attr("gen")

# First meaning: translations (e.g. German, English)
xml_find_all(first_entry, ".//meaning/targetLanguage[@lang='de']/translation") |> xml_text()
xml_find_all(first_entry, ".//meaning/targetLanguage[@lang='en']/translation") |> xml_text()

# Number of examples in the first meaning
xml_find_first(first_entry, ".//meaning/examples") |> xml_attr("count")
```

## Exploring the examples section {.scrollable}

Each **meaning** can contain an **`<examples>`** block with one or more **`<example>`** nodes. Each `<example>` has at least one **`<text>`** element whose children are **`<word>`** and **`<inflectedHeadword>`** (and optionally **`<attribute>`**), in order — that is the Luxembourgish example sentence. Below we **collect** several example sentences from **random entries** and put them into a **tibble** so you can work with them like any table (filter, count, join, etc.).

**Helper:** turn one `<example>` node into the sentence string (first `<text>` only):

```{r}
#| label: lod-get-sentence

# Sentence text from one <example>: first <text>, then concatenate child elements (word, inflectedHeadword, etc.)
get_sentence <- function(ex) {
  text_node <- xml_find_first(ex, ".//text")
  if (inherits(text_node, "xml_missing")) return(NA_character_)
  tokens <- xml_find_all(text_node, "./*")
  paste(xml_text(tokens), collapse = " ")
}
```

**Collect examples and build a tibble:** sample a number of entries, take up to two examples per entry, then combine lemma and sentence into a data frame.

```{r}
#| label: lod-examples-tibble
#| message: false

library(tibble)

all_entries <- xml_find_all(lod_xml, "//entry")
set.seed(123)
n_entries  <- 30
sampled    <- sample(length(all_entries), min(n_entries, length(all_entries)))
rows       <- list()

for (i in seq_along(sampled)) {
  entry  <- all_entries[[sampled[i]]]
  lemma  <- xml_find_first(entry, ".//lemma") |> xml_text()
  exs    <- xml_find_all(entry, ".//meaning/examples/example")
  # Take at most 2 examples per entry
  exs    <- exs[1:min(2, length(exs))]
  for (ex in exs) {
    sent <- get_sentence(ex)
    if (!is.na(sent) && nchar(sent) > 0)
      rows[[length(rows) + 1]] <- list(lemma = lemma, sentence = sent)
  }
}

# One row per example sentence
lod_examples <- tibble(
  lemma    = vapply(rows, function(r) r$lemma,    character(1)),
  sentence = vapply(rows, function(r) r$sentence, character(1))
)

lod_examples
```

You can now use **dplyr** or **tidyr** on `lod_examples` (e.g. count sentences per lemma, filter by length, or join with other data). The same idea scales to all entries if you loop over every entry and every example and bind the rows.

## Summary of the LOD XML structure {.scrollable}

| Level | Element | Role |
|------------------------|------------------------|------------------------|
| Root | `lod` | Container for all entries. |
| Entry | `entry` (id) | One headword: lemma, ipa, microStructure. |
| Micro | `microStructure` | partOfSpeech, inflection(s), grammaticalUnit(s). |
| Unit | `grammaticalUnit` | Groups one or more meanings. |
| Meaning | `meaning` (id, video) | Sense: number, targetLanguage (translation, semanticClarifier), examples. |
| Example | `example` | text (word, inflectedHeadword, optional attribute), optional gloss. |

This structure lets you extract lemmas, translations, parts of speech, inflections, and example sentences (with glosses) for lexical and corpus-oriented analyses in R.

# Hands-on exercises

1.  **Typo.** Why does this fail? (Look very carefully at the object name.)

    ``` r
    my_variable <- 10
    my_varıable
    ```

2.  **Vectors and strings.** In a new code chunk:

    -   Create a character vector `test_words` with at least three words (e.g. `c("word1", "word2", "word3")`).
    -   Use `length(test_words)` and `nchar(test_words)` and interpret the result.

3.  **Comparisons.** Run these and say what they return (TRUE/FALSE) and why:

    ``` r
    "dog" == "dog"
    "dog" != "cat"
    nchar("linguistics") > 5
    ```

4.  **Data import and inspection.** Create a short CSV string with columns `word` and `count` (e.g. three rows). Read it with `read_csv()`, then use `glimpse()` and `$` to inspect the `count` column and compute `mean(count)`.

5.  **Style.** Restyle this into a clear, multi-line pipe (spaces, line breaks, indentation):

    ``` r
    word_data|>filter(freq>3)|>arrange(freq)|>select(word,freq)
    ```

# 
