---
title: "Session 3 — Corpus analysis with tidytext and quanteda"
subtitle: "Language Analytics in R"
author: "Peter Gilles"
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    code-fold: false
editor: visual
---

## Learning objectives

By the end of this session you will:

-   Understand the difference between **tidy text** (one token per row) and **corpus/dfm** workflows.
-   Use **tidytext** to tokenize text and compute word frequencies.
-   Use **quanteda** to build a corpus and a document-term matrix (`dfm`).
-   Compare outputs from tidytext and quanteda on the same corpus.
-   Apply these workflows to linguistic/corpus questions.

# Tidy text vs corpus objects

-   **tidytext**: one token per row (tibbles + dplyr pipeline).
-   **quanteda**: corpus objects and sparse matrices (`dfm`) for text mining.
-   Why both are useful in language analytics.

# Working with Gutenberg text (tidytext)

The **tidytext** package treats text as [one token per row](https://www.tidytextmining.com/tidytext.html).

```{r}
#| label: tidytext-example
#| message: false

library(tidyverse)
library(tidytext)
library(gutenbergr)

# Download one book by Gutenberg ID (35 = The Time Machine)
gutenberg_books <- gutenberg_download(c(35), meta_fields = "title")
head(gutenberg_books)

# Tokenize text to one-word-per-row
tidy_tokens <- gutenberg_books |>
  unnest_tokens(output = word, input = text)

head(tidy_tokens, 12)
```

```{r}
#| label: tidytext-count
#| message: false

# Remove stop words and count
tidy_tokens |>
  anti_join(stop_words, by = "word") |>
  count(word, sort = TRUE) |>
  head(15)
```

# Working with corpus objects (quanteda)

Use the same Gutenberg data to build a corpus and a document-term matrix.

```{r}
#| label: quanteda-example
#| message: false

library(quanteda)
library(quanteda.textstats)

corp <- corpus(gutenberg_books, text_field = "text")
summary(corp, n = 5)
ndoc(corp)
ntoken(corp)
```

```{r}
#| label: quanteda-dfm

doc_term_matrix <- corp |>
  tokens(remove_punct = TRUE) |>
  dfm()

dim(doc_term_matrix)
textstat_frequency(doc_term_matrix) |>
  head(10)
```

# Comparing outputs

-   **tidytext** output: token table (easy with dplyr and joins, e.g. sentiment lexicons).
-   **quanteda** output: corpus + `dfm` (efficient for matrix-based text mining/classification).
-   Same corpus, different data structures and downstream strengths.

# Hands-on exercises

1.  Download a Gutenberg text (`gutenberg_download()`), tokenize with `unnest_tokens()`, remove stop words, and list top 20 words.
2.  Build a quanteda corpus and `dfm` for the same text; inspect `ndoc()`, `ntoken()`, and top terms.
3.  Write 2–3 sentences comparing tidytext and quanteda outputs for your text.

# Summary and further reading

-   [Text Mining with R — tidy text format](https://www.tidytextmining.com/tidytext)
-   [gutenbergr text-mining vignette](https://docs.ropensci.org/gutenbergr/articles/text-mining.html)
-   [A&S500 class notes (text + quanteda)](https://jofrhwld.github.io/AandS500_2023/)
