---
title: "Session 2 — Data wrangling"
subtitle: "Language Analytics in R"
author: "Peter Gilles"
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    code-fold: false
editor: visual
---

## Learning objectives

By the end of this session you will:

-   Load, inspect, and aggregate CSV data with tidyverse tools.
-   Import and inspect complex XML dictionary data (LOD).
-   Use XPath selectors to extract linguistic information from XML.
-   Build a tibble of example sentences from XML nodes.

# Working with text data and corpora (data import)

For language analytics we usually work with **tables** (e.g. CSV or Excel) or **text files**.

## Reading a CSV file {.scrollable}

Use `read_csv()` from the **readr** package (included in tidyverse) to read comma-separated files. Example: population by commune (Luxembourg), file `rnrpp-population-commune.csv` (same folder as this document, or adjust the path if needed).

```{r}
#| label: read-csv
#| echo: false

library(tidyverse)

# Path: same folder as this .qmd, or "sessions/rnrpp-population-commune.csv" from project root
pop_path <- "rnrpp-population-commune.csv"
pop <- read_csv(pop_path, show_col_types = FALSE)
pop
```

## Inspecting the dataset {.scrollable}

-   **Print** the data frame: type its name.
-   **View(pop)**: opens a spreadsheet-like viewer in RStudio.
-   **glimpse()**: compact view of columns and types.

```{r}
#| label: inspect-data

glimpse(pop)
nrow(pop)
ncol(pop)
names(pop)
head(pop, 5)
```

## Accessing columns {.scrollable}

Use `$` to get one column as a vector (e.g. to compute totals or means):

```{r}
#| label: access-columns

pop$COMMUNE_NOM
pop$FEMMES_MAJEURES
# Mean population per commune (total = sum of the four count columns)
mean(pop$FEMMES_MINEURES + pop$HOMMES_MINEURS + pop$FEMMES_MAJEURES + pop$HOMMES_MAJEURS)
```

## Aggregating the data {.scrollable}

With **dplyr** you can **aggregate** by groups (e.g. totals, means). Example: add a total population per commune, then show the top 10 communes by total population; or summarize across all rows (e.g. national totals by category).

```{r}
#| label: aggregate-pop

# Total population per commune (using the pipe)
pop |>
  mutate(
    total = FEMMES_MINEURES + HOMMES_MINEURS + FEMMES_MAJEURES + HOMMES_MAJEURS
  ) |>
  arrange(desc(total)) |>
  select(COMMUNE_NOM, total) |>
  head(10)
```

```{r}
#| label: aggregate-summarize

# One-row summary: national totals by category (sum across all communes)
pop |>
  summarise(
    femmes_mineures = sum(FEMMES_MINEURES),
    hommes_mineurs  = sum(HOMMES_MINEURS),
    femmes_majeures = sum(FEMMES_MAJEURES),
    hommes_majeurs  = sum(HOMMES_MAJEURS),
    total          = sum(FEMMES_MINEURES + HOMMES_MINEURS + FEMMES_MAJEURES + HOMMES_MAJEURS)
  )
```

# Import LOD dictionary data (XML)

## Getting the file {.scrollable}

The LOD linguistic dataset is published as **open data** by the Luxembourg government (Zenter fir d'Lëtzebuerger Sprooch) under **CC0** ([LOD – Linguistesch Daten (resource)](https://data.public.lu/fr/datasets/letzebuerger-online-dictionnaire-lod-linguistesch-daten/#/resources/7aad8691-006f-43b0-8d65-e47e8d51696b)). Download the zip, extract it, and place the XML file (`new_lod-art.xml`, ~74 MB) in a folder in your project (e.g. `data/`).

**Install once:** `install.packages("xml2")`. Then load with `library(xml2)`.

## Reading the XML file {.scrollable}

```{r}
#| label: lod-read-xml
#| message: false

library(xml2)

# Path to the XML file (e.g. in data/; adjust if your file is elsewhere)
lod_xml_path <- "new_lod-art.xml"

# Read XML from the folder
lod_xml <- read_xml(lod_xml_path)

# Root element
xml_name(lod_xml)
xml_children(lod_xml) |> length()
```

## Structure of the LOD XML {.scrollable}

The document has a single root element `<lod>`, which contains many **`<entry>`** elements. Each entry has a lemma, pronunciation, grammatical structure, meanings, translations, and examples.

## How XPath works

**XPath** selects nodes in the XML tree. In R, use `xml_find_first()` (one match) and `xml_find_all()` (all matches).

| XPath | Meaning | Example |
|---|---|---|
| `//entry` | any `entry` in document | all entries |
| `.//lemma` | lemma under current node | current entry lemma |
| `.//meaning/examples/example` | examples under meanings | all examples for an entry |
| `targetLanguage[@lang='en']` | attribute filter | English translation nodes |

## Inspecting one entry in R {.scrollable}

```{r}
#| label: lod-structure

first_entry <- xml_find_first(lod_xml, "//entry")
xml_find_first(first_entry, ".//lemma") |> xml_text()
xml_find_first(first_entry, ".//ipa")   |> xml_text()
xml_find_all(first_entry, ".//meaning/targetLanguage[@lang='en']/translation") |> xml_text()
```

## Exploring the examples section {.scrollable}

```{r}
#| label: lod-get-sentence

get_sentence <- function(ex) {
  text_node <- xml_find_first(ex, ".//text")
  if (inherits(text_node, "xml_missing")) return(NA_character_)
  tokens <- xml_find_all(text_node, "./*")
  paste(xml_text(tokens), collapse = " ")
}
```

```{r}
#| label: lod-examples-tibble
#| message: false

library(tibble)

all_entries <- xml_find_all(lod_xml, "//entry")
set.seed(123)
sampled <- sample(length(all_entries), min(30, length(all_entries)))
rows <- list()

for (i in seq_along(sampled)) {
  entry <- all_entries[[sampled[i]]]
  lemma <- xml_find_first(entry, ".//lemma") |> xml_text()
  exs <- xml_find_all(entry, ".//meaning/examples/example")
  exs <- exs[1:min(2, length(exs))]
  for (ex in exs) {
    sent <- get_sentence(ex)
    if (!is.na(sent) && nchar(sent) > 0) {
      rows[[length(rows) + 1]] <- list(lemma = lemma, sentence = sent)
    }
  }
}

lod_examples <- tibble(
  lemma = vapply(rows, function(r) r$lemma, character(1)),
  sentence = vapply(rows, function(r) r$sentence, character(1))
)

lod_examples
```

## Hands-on exercises

1. Read `rnrpp-population-commune.csv` and compute total population per commune.
2. For LOD XML, extract lemma and first English translation for 10 entries.
3. Build a tibble of random example sentences (lemma + sentence) and count rows per lemma.
