---
title: "Session 3 — Introduction to quanteda"
subtitle: "Language Analytics in R"
author: "Peter Gilles"
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    code-fold: false
editor: visual
---

## Learning objectives

By the end of this session you will:

-   Understand core **quanteda** objects (`corpus`, `tokens`, `dfm`, `docvars`).
-   Build and inspect corpora from both package data and Project Gutenberg data.
-   Use `kwic()` for context-based exploration and `topfeatures()` for frequency analysis.
-   Group documents via document variables and apply a simple dictionary analysis.
-   Run an end-to-end mini analysis workflow with **quanteda**.

# Setup

In this session we combine:

-   **quanteda** for corpus objects and matrix-based analysis,
-   **quanteda.textstats/textplots** for basic statistics and visual output,
-   **gutenbergr** to fetch open literary data.

Run this chunk once at the beginning. If a package is missing, install it with `install.packages()`.

```{r}
#| label: setup
#| message: false

library(quanteda)
library(quanteda.textstats)
library(quanteda.textplots)
library(gutenbergr)
library(dplyr)
library(ggplot2)
```

# Introduction to quanteda

**quanteda** is a text analysis framework for R that is optimized for corpus workflows:

-   keep original texts in a `corpus`,
-   transform texts into `tokens`,
-   represent token counts in a sparse matrix (`dfm`),
-   combine textual data with metadata (`docvars`) for linguistic and social analysis.

In short: quanteda is useful when you want transparent preprocessing, efficient counting, and reproducible corpus-based analysis.

## 1) Build a corpus from built-in quanteda data

`data_corpus_inaugural` is bundled with quanteda and contains U.S. inaugural speeches with document metadata (`docvars`) such as `Year`, `President`, and `Party`.

In quanteda, **docvars** are document-level variables attached to each text (comparable to columns in a metadata table).\
They are crucial because they let us:

-   filter documents (`corpus_subset(corp, Year >= 1993)`),
-   compare groups (`dfm_group(..., groups = Party)`),
-   model textual outcomes with external information.

The goal of this first step is to inspect the corpus structure before any preprocessing:

-   how many documents we have,
-   how large they are,
-   which metadata variables are available for grouping and filtering.

```{r}
#| label: corpus-built-in
#| message: false

corp_inaug <- data_corpus_inaugural
summary(corp_inaug, n = 6)
head(docvars(corp_inaug))
```

To get a direct glimpse of corpus content, inspect one or two documents as plain character text.\
For fast corpus statistics, use:

-   `summary(corpus_obj)` for document-level overview,
-   `ndoc(corpus_obj)` for number of documents,
-   `ntoken(corpus_obj)` for total token count,
-   `nfeat(dfm_obj)` once a dfm exists (number of features/columns).

```{r}
#| label: corpus-glimpse-and-stats

# A readable text glimpse with explicit labels
cat("First document name:", docnames(corp_inaug)[1], "\n")
cat(as.character(corp_inaug)[1], "\n\n")

cat("Second document name:", docnames(corp_inaug)[2], "\n")
cat(as.character(corp_inaug)[2], "\n\n")

# Core corpus statistics with labels
cat("Number of documents (ndoc):", ndoc(corp_inaug), "\n")
cat("Total number of tokens (ntoken):", ntoken(corp_inaug), "\n")
```

You can also build a corpus from your own text file.\
The next chunk reads `Bibel.txt` from the same folder as this session file, creates a one-document corpus, and adds docvars manually.

```{r}
#| label: own-file-corpus
#| message: false

bibel_path <- file.path("Bibel.txt")

bibel_text <- readLines(bibel_path, encoding = "UTF-8", warn = FALSE) |>
  paste(collapse = "\n")

corp_bibel <- corpus(
  x = bibel_text,
  docnames = "Bibel"
)
corp_bibel

# Example docvars: these can be changed/adapted
docvars(corp_bibel, "language") <- "lb"
docvars(corp_bibel, "genre") <- "religious_text"

summary(corp_bibel)
```

## 2) Subset by document variables

Now we narrow the corpus to recent speeches (`Year >= 1993`). This mirrors a common workflow: define a specific analytical window before tokenization.

```{r}
#| label: corpus-subset

corp_recent <- corpus_subset(corp_inaug, Year >= 1993)
summary(corp_recent, n = 6)
```

## 3) Tokenize and preprocess

This is the central preprocessing step:

-   `tokens()` splits text into tokens,
-   `remove_punct` and `remove_numbers` reduce noise,
-   `tokens_tolower()` normalizes case,
-   `tokens_remove(stopwords("en"))` removes frequent function words.

We keep the original corpus unchanged and store transformed text in a new object (`toks_recent`).

```{r}
#| label: tokens-preprocess

toks_recent <- tokens(
  corp_recent,
  remove_punct = TRUE,
  remove_numbers = TRUE
) |>
  tokens_tolower() |>
  tokens_remove(stopwords("en"))

toks_recent
```

## 4) Keywords in context (KWIC)

`kwic()` lets us inspect how a term is used in context, not just how often it occurs. This is useful for linguistic interpretation because it prevents over-interpreting frequency counts alone.

```{r}
#| label: kwic-example

kwic(toks_recent, pattern = "freedom") |>
  head(12)
```

Further example with wildcard patterns (`valuetype = "glob"`):

```{r}
#| label: kwic-wildcard

kwic(toks_recent, pattern = "democra*", valuetype = "glob") |>
  head(12)
```

## 5) Create and inspect a dfm

A **document-feature matrix** (`dfm`) is a sparse matrix with:

-   rows = documents,
-   columns = features (tokens),
-   cells = feature counts.

Why this matrix matters:

-   it transforms text into a numeric format required for statistics and machine learning,
-   it is sparse and memory-efficient (important for large corpora),
-   it allows direct operations such as feature selection, weighting, grouping, and modeling.

Typical use cases include keyness analysis, topic/style comparison, supervised classification, and dimensional reduction.

```{r}
#| label: dfm-example

# Build document-feature matrix from cleaned tokens
dfm_recent <- dfm(toks_recent)

# Inspect matrix dimensions and sparsity
dfm_recent

# Show the 20 most frequent features across all documents
topfeatures(dfm_recent, 20)
```

Using the dfm, we can also easily create word clouds.

```{r}
textplot_wordcloud(dfm_recent, min_count = 20, random_order = FALSE,
                   rotation = .25)
```

## 6) Group by docvars (Party) and compare

With `dfm_group()`, we aggregate all documents that share the same value of a document variable (`Party`). This moves from document-level comparison to group-level comparison.

```{r}
#| label: dfm-group

dfm_party <- dfm_recent |>
  dfm_group(groups = Party)

dfm_party
topfeatures(dfm_party, 20)
```

## 7) Simple dictionary analysis

A dictionary maps many individual words to a smaller set of analytical categories.\
Here we define three thematic categories and then count category matches per document.

Pedagogically, this demonstrates how to move from raw lexical frequencies to theory-driven categories.

```{r}
#| label: dictionary-analysis

# Define a small dictionary with 3 thematic categories
issue_dict <- dictionary(
  list(
    economy = c("economy", "jobs", "work", "business", "growth"),
    security = c("security", "terror", "terrorism", "war", "threat"),
    democracy = c("democracy", "freedom", "rights", "constitution", "people")
  )
)

# Map matching tokens to dictionary categories, then count by document
dfm_dict <- toks_recent |>
  tokens_lookup(dictionary = issue_dict) |>
  dfm()

# Inspect category counts by document
dfm_dict

# Summarize most frequent dictionary categories in the corpus
textstat_frequency(dfm_dict, n = 15)
```

Interpretation tip: a high value for a category means that words from this category appear frequently in a document.\
Because categories have different dictionary sizes and texts have different lengths, compare results carefully (relative frequencies can be useful in advanced analysis).

# Quanteda with Gutenberg data

We now repeat the workflow on an external corpus from Project Gutenberg, using German literary texts.

For these Faust texts, a direct UTF-8 text import is more reliable than `gutenberg_download()` and avoids encoding artifacts that can split words into fragments.

```{r}
#| label: gutenberg-data
#| message: false

load_faust_lines <- function(gutenberg_id, title, author) {
  url <- sprintf(
    "https://www.gutenberg.org/cache/epub/%s/pg%s.txt",
    gutenberg_id, gutenberg_id
  )
  con <- url(url)
  on.exit(close(con), add = TRUE)

  lines <- readLines(con, warn = FALSE, encoding = "UTF-8")
  lines <- gutenbergr::gutenberg_strip(lines)

  tibble(
    gutenberg_id = gutenberg_id,
    title = title,
    author = author,
    text = lines
  )
}

books <- bind_rows(
  load_faust_lines(2229, "Faust: Der Tragödie erster Teil", "Johann Wolfgang von Goethe"),
  load_faust_lines(2230, "Faust: Der Tragödie zweiter Teil", "Johann Wolfgang von Goethe")
)

books |>
  distinct(gutenberg_id, title, author)

head(books)

# Collapse line-level Gutenberg rows into one document per title
faust_docs <- books |> 
  summarise(text = paste(text, collapse = "\n"), .by = title)

head(faust_docs)
```

```{r}
#| label: gutenberg-corpus

corp_faust <- corpus(faust_docs, text_field = "text")
summary(corp_faust, n = 10)
```

Again, we tokenize, normalize, remove German stopwords, and build a `dfm`.

```{r}
#| label: gutenberg-dfm

# Tokenize and clean German texts
toks_gut <- corp_faust |>
  tokens(remove_punct = TRUE, remove_numbers = TRUE) |>
  tokens_tolower() |>
  tokens_remove(stopwords("de"))

dfm_gut <-  dfm(toks_gut)

# Compact structural overview
cat("Documents:", ndoc(dfm_gut), "\n")
cat("Features:", nfeat(dfm_gut), "\n")

# Most frequent features
top_25_gut <- topfeatures(dfm_gut, 25)
top_25_gut
```

## Exercise 1: Moby-Dick workflow

In this exercise, you apply the full quanteda pipeline to one classic text.

### Task

1.  Load *Moby-Dick* from Gutenberg (`ID = 2701`).
2.  Build a quanteda corpus from the downloaded text.
3.  Tokenize the corpus (remove stopwords, punctuation and numbers, lowercase).
4.  Run some `kwic()` searches with a flexible `valuetype` argument (`"fixed"`, `"glob"`, `"regex"`), e.g.:
    -   an exact search (`"whale"`),
    -   a wildcard search (`"whal*"`),
    -   a regex search (`"whal(e|es|ing)"`).

::: {.callout-tip collapse="true"}
## Possible solution — Exercise 1

```{r}
#| label: exercise-mobydick-data
#| message: false

# 1) Load Moby-Dick from Gutenberg
moby_raw <- gutenberg_download(2701, meta_fields = c("title", "author"))

# Keep only key metadata rows for inspection
moby_raw |>
  distinct(gutenberg_id, title, author)
```

```{r}
#| label: exercise-mobydick-corpus

# 2) Build corpus
corp_moby <- corpus(moby_raw, text_field = "text")
summary(corp_moby, n = 5)
```

```{r}
#| label: exercise-mobydick-tokens

# 3) Tokenize and basic preprocessing
toks_moby <- tokens(
  corp_moby,
  remove_punct = TRUE,
  remove_numbers = TRUE
) |>
  tokens_tolower() |>
  tokens_remove(stopwords("en"))

toks_moby
```

```{r}
#| label: exercise-mobydick-kwic-function

# Exact search
kwic(toks_moby, pattern = "whale", valuetype = "fixed", n = 8)

# Wildcard search
kwic(toks_moby, pattern = "whal*", valuetype = "glob", n = 8)

# Regex search
kwic(toks_moby, pattern = "whal(e|es|ing)", valuetype = "regex", n = 8)
```
:::

## Exercise 2: Lexical dispersion plot with `textplot_xray()`

In this exercise, you create a lexical dispersion plot for the **Faust** corpus (*Faust I* + *Faust II*) similar to the quanteda plotting example:

-   [quanteda textual data visualization example](https://quanteda.io/articles/pkgdown/examples/plotting.html)

### Task

1.  Open the help page for `textplot_xray()` and review:
    -   required input object(s),

    -   how multiple `kwic()` objects are passed,

    -   main plotting options.
2.  Using your Faust Gutenberg data, first collapse line-level rows to one text per title, then create a dispersion plot for, e.g. (or other words):
    -   `"gretchen"`
    -   `"teufel"`

::: {.callout-tip collapse="true"}
## Possible solution — Exercise 2 (click to expand)

```{r}
#| label: exercise-faust-xray-solution

corp_faust <- corpus(faust_docs, text_field = "text")

# Build tokens from the collapsed Faust corpus
toks_faust <- tokens(corp_faust, remove_punct = TRUE) |>
  tokens_tolower()

# Create KWIC objects for both search terms
kwic_1 <- kwic(toks_faust, pattern = "himmel")
kwic_2 <- kwic(toks_faust, pattern = "hölle")

# Plot lexical dispersion ("x-ray")
textplot_xray(
  kwic_1,
  kwic_2
)
```
:::

# Introduction to ggplot2

## Generic introduction

`ggplot2` is based on the **grammar of graphics**.\
The core idea is that every plot is built from the same components:

-   **data** (the dataset),
-   **aesthetics** (`aes()`: which variable maps to x, y, color, etc.),
-   **geometry** (`geom_*()`: bars, points, lines, polygons),
-   **optional layers** (labels, scales, themes, facets).

The canonical structure is:

```{r}
#| eval: false

ggplot(data = ..., aes(x = ..., y = ...)) +
  geom_...() +
  labs(...) +
  theme_minimal()
```

Note that the components in ggplot2 are connected with the + sign.

## First plots

The dataset `mpg` is included in ggplot2.\
Each row is a car model observation, with variables such as:

-   `class`: vehicle type (SUV, compact, etc.),
-   `manufacturer`: car brand,
-   `displ`: engine displacement (liters),
-   `hwy`: highway fuel economy (miles per gallon).

It is a good teaching dataset because it contains both categorical and numeric variables. When ggplot2 is loaded, the dataset is immediately available as `mpg`.

```{r}
#| label: ggplot-easy-1

ggplot(mpg, aes(x = class)) +
  geom_bar()
```

```{r}
#| label: ggplot-easy-2

ggplot(mpg, aes(x = class)) +
  geom_bar(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Number of cars by class",
    x = "Car class",
    y = "Count"
  ) +
  theme_minimal()
```

## Bar charts

`geom_bar()` counts rows automatically.\
`geom_col()` expects a pre-computed y-value.

When to use which:

-   Use `geom_bar()` if your question is "how many observations are in each category?".
-   Use `geom_col()` if you already calculated a statistic (mean, median, sum) per category.

In the first example, ggplot2 performs the counting for us.

```{r}
#| label: ggplot-barchart-count

ggplot(mpg, aes(x = manufacturer)) +
  geom_bar(fill = "darkorange") +
  coord_flip() +
  labs(
    title = "Observations by manufacturer",
    x = "Manufacturer",
    y = "Count"
  ) +
  theme_minimal()
```

```{r}
#| label: ggplot-barchart-mean

# First, compute mean highway fuel economy by car class
mean_hwy_by_class <- mpg |>
  summarise(mean_hwy = mean(hwy), .by = class)

# Then plot those pre-computed values
ggplot(mean_hwy_by_class, aes(x = class, y = mean_hwy)) +
  geom_col(fill = "seagreen") +
  coord_flip() +
  labs(
    title = "Mean highway MPG by class",
    x = "Class",
    y = "Mean highway MPG"
  ) +
  theme_minimal()
```

## Boxplots and violin plots with `iris`

The `iris` dataset is a classic built-in dataset with 150 flowers from 3 species:

-   `Sepal.Length`, `Sepal.Width`, `Petal.Length`, `Petal.Width` (numeric measurements),
-   `Species` (categorical label: setosa, versicolor, virginica).

A **boxplot** summarizes distribution (median, quartiles, outliers).\
A **violin plot** shows the full density shape of the distribution.

```{r}
#| label: ggplot-boxplot-iris

ggplot(iris, aes(x = Species, y = Sepal.Length, fill = Species)) +
  geom_boxplot(alpha = 0.8) +
  labs(
    title = "Sepal length by species (boxplot)",
    x = "Species",
    y = "Sepal length"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

```{r}
#| label: ggplot-violin-iris

ggplot(iris, aes(x = Species, y = Sepal.Length, fill = Species)) +
  geom_violin(alpha = 0.6, color = "grey20") +
  geom_boxplot(width = 0.12, fill = "white", outlier.size = 0.8) +
  labs(
    title = "Sepal length by species (violin + boxplot)",
    x = "Species",
    y = "Sepal length"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

Color options you can discuss with students:

-   Set one fixed color: `fill = "steelblue"` in `geom_*()`.
-   Map color/fill to a variable: `aes(fill = Species)`.
-   Use manual palettes: `scale_fill_manual(values = c(...))`.
-   Use ColorBrewer palettes: `scale_fill_brewer(palette = "Set2")`.

## Scatter plots

```{r}
#| label: ggplot-scatter-basic

ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(alpha = 0.7) +
  labs(
    title = "Engine displacement and highway MPG",
    x = "Engine displacement (liters)",
    y = "Highway MPG"
  ) +
  theme_minimal()
```

```{r}
#| label: ggplot-scatter-colored

# Linear model and correlation (r) for displ ~ hwy
lm_mpg <- lm(hwy ~ displ, data = mpg)
summary(lm_mpg)

r_displ_hwy <- cor(mpg$displ, mpg$hwy, use = "complete.obs")
cat("Pearson r (displ, hwy):", round(r_displ_hwy, 3), "\n")

ggplot(mpg, aes(x = displ, y = hwy, color = class)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.8) +
  labs(
    title = "Displacement vs highway MPG by class",
    x = "Engine displacement (liters)",
    y = "Highway MPG",
    color = "Class"
  ) +
  theme_minimal()
```

## Maps with Luxembourg examples

For a lightweight first map, we use polygon data from `map_data("world")` and zoom into Luxembourg.

```{r}
#| label: ggplot-map-lux-region
#| fig-width: 11
#| fig-height: 7
#| out-width: "100%"

world_map <- map_data("world")

ggplot(world_map, aes(x = long, y = lat, group = group)) +
  geom_polygon(fill = "grey90", color = "white", linewidth = 0.2) +
  coord_quickmap(xlim = c(4.5, 7.0), ylim = c(49.2, 50.4)) +
  labs(
    title = "Luxembourg in regional context",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_minimal()
```

```{r}
#| label: ggplot-map-lux-highlight
#| fig-width: 12
#| fig-height: 18
#| out-width: "100%"
#| fig.show: "hold"

world_map <- map_data("world")
communes_df <- readRDS("communes_df.RDS")
centroids_label <- readRDS("centroids_label.RDS")

lux_map <- world_map |>
  filter(region == "Luxembourg")

# Step 1: regional background
p1 <- ggplot() +
  geom_polygon(
    data = world_map,
    aes(x = long, y = lat, group = group),
    fill = "grey92", color = "white", linewidth = 0.2
  ) +
  coord_quickmap(xlim = c(5.5, 6.7), ylim = c(49.2, 50.4)) +
  labs(
    title = "Step 1: Regional background",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_minimal()
print(p1)

# Step 2: add commune polygons
p2 <- p1 +
  # Layer 2: commune polygons inside Luxembourg
  geom_polygon(
    data = communes_df,
    aes(x = long, y = lat, group = group),
    fill = "#fdebd0", color = "#d35400", linewidth = 0.15, alpha = 0.8
  ) +
  labs(title = "Step 2: + Commune polygons")
print(p2)

# Step 3: add Luxembourg country outline
p3 <- p2 +
  # Layer 3: Luxembourg country outline
  geom_polygon(
    data = lux_map,
    aes(x = long, y = lat, group = group),
    fill = NA, color = "black", linewidth = 0.5
  ) +
  labs(title = "Step 3: + Country outline")
print(p3)

# Step 4: add centroid points for important places
p4 <- p3 +
  # Layer 4: important city points from centroid file
  geom_point(
    data = centroids_label,
    aes(x = long, y = lat),
    color = "navy", size = 1.8
  ) +
  labs(title = "Step 4: + Important place points")
print(p4)

# Step 5: add city labels (final map)
p5 <- p4 +
  # Layer 5: city labels
  geom_text(
    data = centroids_label,
    aes(x = long, y = lat, label = label),
    size = 2.8, nudge_y = 0.015, color = "navy"
  ) +
  labs(
    title = "Step 5: Final map",
    subtitle = "Communes (polygons) + centroid city labels",
    x = "Longitude",
    y = "Latitude"
  )
print(p5)
```

You can also combine the five map states into one overview grid:

```{r}
#| label: ggplot-map-layers-grid
#| fig-width: 13
#| fig-height: 10
#| out-width: "100%"

gridExtra::grid.arrange(p1, p2, p3, p4, p5, ncol = 5)
```

Step-by-step map logic:

1.  Load base map polygons (`map_data("world")`) and zoom to Luxembourg.
2.  Add commune polygons from `communes_df.RDS` as an internal administrative layer.
3.  Draw Luxembourg border on top for a clear country outline.
4.  Add centroid points from `centroids_label.RDS`.
5.  Add text labels for important places (`label` column).

## Population per commune (choropleth)

This map visualizes total population (`total_population`) by commune.

How it works:

1.  Load commune population table from `Gemengen_Statistiken.qs`.
2.  Join population values to commune polygons via commune name (`id` vs `COMMUNE_NOM`).
3.  Use `fill = total_population` to encode population with a color gradient.
4.  Optionally add commune labels at centroids for orientation.

```{r}
#| label: ggplot-map-population-communes
#| fig-width: 14
#| fig-height: 10
#| out-width: "100%"

communes_df <- readRDS("communes_df.RDS")
centroids_label <- readRDS("centroids_label.RDS")

# Preferred source: Gemengen_Statistiken.qs
pop_totals <- readRDS("Gemengen_Statistiken.RDS")

# Join population table to polygon coordinates
communes_pop_map <- communes_df |>
  left_join(
    pop_totals |>
      select(Gemeng, total_population = Awunnerzuel),
    by = c("id" = "Gemeng")
  )

ggplot(communes_pop_map, aes(x = long, y = lat, group = group, fill = total_population)) +
  geom_polygon(color = "white", linewidth = 0.15) +
  geom_text(
    data = centroids_label,
    aes(x = long, y = lat, label = label),
    inherit.aes = FALSE,
    size = 2.4,
    color = "grey20"
  ) +
  coord_quickmap(xlim = c(5.5, 6.7), ylim = c(49.2, 50.2)) +
  labs(
    title = "Luxembourg: total population by commune",
    subtitle = "Fill color = total_population",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_minimal()
```

Interpretation:

-   Darker communes have lower population totals.
-   Luxembourg City should stand out clearly as the most populous commune.

## Exercises

### Exercise A (`mpg`): from raw data to summary plots

**Goal:** practice both direct plotting and plotting after data restructuring.

Tasks:

1.  Create a simple count plot of `class` with `geom_bar()`.
2.  Build a summary table with:
    -   `mean_hwy` (mean of `hwy`) by `class`
    -   `sum_cty` (sum of `cty`) by `class`
3.  Plot both summaries with `geom_col()` (two separate plots).
4.  Briefly interpret one pattern (2-3 lines).

::: {.callout-tip collapse="true"}
## Possible solution — Exercise A (click to expand)

```{r}
# A1: direct count plot from raw rows
ggplot(mpg, aes(x = class)) +
  geom_bar(fill = "steelblue") +
  coord_flip() +
  theme_minimal()

# A2: restructure data (mean and sum by class)
mpg_summary <- mpg |>
  summarise(
    mean_hwy = mean(hwy, na.rm = TRUE),
    sum_cty = sum(cty, na.rm = TRUE),
    .by = class
  )

# A3a: plot mean_hwy
ggplot(mpg_summary, aes(x = class, y = mean_hwy)) +
  geom_col(fill = "seagreen") +
  coord_flip() +
  labs(title = "Mean highway MPG by class", x = "Class", y = "Mean hwy") +
  theme_minimal()

# A3b: plot sum_cty
ggplot(mpg_summary, aes(x = class, y = sum_cty)) +
  geom_col(fill = "darkorange") +
  coord_flip() +
  labs(title = "Sum of city MPG values by class", x = "Class", y = "Sum cty") +
  theme_minimal()
```
:::

### Exercise B (`iris`): distribution plots and grouping summaries

**Goal:** compare distributions and aggregated means across species.

Tasks:

1.  Create a boxplot of `Sepal.Length` by `Species`.
2.  Create a violin plot of `Sepal.Length` by `Species` and overlay a narrow boxplot.
3.  Build a summary table with mean values of `Sepal.Length` and `Petal.Length` by species.
4.  Plot the mean `Sepal.Length` by species as `geom_col()`.

::: {.callout-tip collapse="true"}
## Possible solution — Exercise B (click to expand)

```{r}
# B1: boxplot
ggplot(iris, aes(x = Species, y = Sepal.Length, fill = Species)) +
  geom_boxplot(alpha = 0.8) +
  theme_minimal() +
  theme(legend.position = "none")

# B2: violin + boxplot overlay
ggplot(iris, aes(x = Species, y = Sepal.Length, fill = Species)) +
  geom_violin(alpha = 0.6, color = "grey20") +
  geom_boxplot(width = 0.12, fill = "white", outlier.size = 0.8) +
  theme_minimal() +
  theme(legend.position = "none")

# B3: restructure with species-level means
iris_summary <- iris |>
  summarise(
    mean_sepal = mean(Sepal.Length, na.rm = TRUE),
    mean_petal = mean(Petal.Length, na.rm = TRUE),
    .by = Species
  )

# B4: plot mean sepal length
ggplot(iris_summary, aes(x = Species, y = mean_sepal, fill = Species)) +
  geom_col() +
  labs(title = "Mean Sepal Length by species", y = "Mean Sepal Length") +
  theme_minimal() +
  theme(legend.position = "none")
```
:::

### Exercise C (population): choropleth with joins and aggregation

**Goal:** practice map joins and pre-aggregation for cartographic input.

Tasks:

1.  Join `communes_df` with `pop_totals` to map `total_population` by commune.
2.  Create the choropleth (`fill = total_population`) with commune labels.
3.  Build a 5-class quantile variable (`pop_quantile`) from `total_population`.
4.  Plot a second choropleth using `fill = pop_quantile` (discrete classes).

::: {.callout-tip collapse="true"}
## Possible solution — Exercise C (click to expand)

```{r}
communes_df <- readRDS("communes_df.RDS")
centroids_label <- readRDS("centroids_label.RDS")
pop_totals <- readRDS("Gemengen_Statistiken.RDS")

# C1: join map polygons with commune population stats
communes_pop_map <- communes_df |>
  left_join(
    pop_totals |>
      select(Gemeng, total_population = Awunnerzuel),
    by = c("id" = "Gemeng")
  )

# C2: continuous choropleth
ggplot(communes_pop_map, aes(x = long, y = lat, group = group, fill = total_population)) +
  geom_polygon(color = "white", linewidth = 0.15) +
  geom_text(
    data = centroids_label,
    aes(x = long, y = lat, label = label),
    inherit.aes = FALSE,
    size = 2.2
  ) +
  coord_quickmap(xlim = c(5.5, 6.7), ylim = c(49.2, 50.2)) +
  labs(title = "Population by commune", fill = "Population") +
  theme_minimal()

# C3: aggregation/restructuring into quantile groups
communes_pop_map <- communes_pop_map |>
  mutate(pop_quantile = dplyr::ntile(total_population, 5))

# C4: discrete choropleth (5 classes)
ggplot(communes_pop_map, aes(x = long, y = lat, group = group, fill = factor(pop_quantile))) +
  geom_polygon(color = "white", linewidth = 0.15) +
  coord_quickmap(xlim = c(5.5, 6.7), ylim = c(49.2, 50.2)) +
  labs(title = "Population quantiles by commune", fill = "Quantile") +
  theme_minimal()
```
:::
