---
title: "Session 2 — Exercise solutions"
subtitle: "Language Analytics in R"
author: "Peter Gilles"
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    code-fold: false
editor: visual
---

## How to use this file

This document contains explained solutions for the exercises in `sessions/session-02.qmd`.

```{r}
#| label: setup
#| message: false

library(tidyverse)
library(tidytext)
library(gutenbergr)
library(xml2)

resolve_path <- function(filename) {
  if (file.exists(filename)) {
    filename
  } else {
    file.path("sessions", filename)
  }
}
```

# CSV exercises — solutions

## Exercise 1

**Task:** Compute the total number of minors per commune and list the top 15.

```{r}
#| label: csv-solution-1

pop <- read_csv(
  resolve_path("rnrpp-population-commune.csv"),
  locale = locale(encoding = "ISO-8859-1"),
  show_col_types = FALSE
)

top_minors <- pop |>
  mutate(minors_total = FEMMES_MINEURES + HOMMES_MINEURS) |>
  arrange(desc(minors_total)) |>
  select(COMMUNE_NOM, minors_total) |>
  slice_head(n = 15)

top_minors
```

Why this works:

-   `mutate()` creates a new variable without changing the raw input columns.
-   `arrange(desc())` sorts from highest to lowest.
-   `slice_head(n = 15)` keeps exactly the requested number of rows.

## Exercise 2

**Task:** Add `female_share` and report top 10 communes.

```{r}
#| label: csv-solution-2

female_share_tbl <- pop |>
  mutate(
    total_population = FEMMES_MINEURES + HOMMES_MINEURS + FEMMES_MAJEURES + HOMMES_MAJEURS,
    female_total = FEMMES_MINEURES + FEMMES_MAJEURES,
    female_share = female_total / total_population
  ) |>
  arrange(desc(female_share)) |>
  select(COMMUNE_NOM, total_population, female_share) |>
  slice_head(n = 10)

female_share_tbl
```

Interpretation note:

-   `female_share` is a proportion between 0 and 1.
-   For reporting, you can add `mutate(female_share_pct = scales::percent(female_share))`.

## Exercise 3

**Task:** Plot a histogram of commune population and describe the shape.

```{r}
#| label: csv-solution-3
#| fig-width: 8
#| fig-height: 5

pop |>
  mutate(total_population = FEMMES_MINEURES + HOMMES_MINEURS + FEMMES_MAJEURES + HOMMES_MAJEURS) |>
  ggplot(aes(x = total_population)) +
  geom_histogram(bins = 25, fill = "steelblue", color = "white") +
  labs(
    title = "Distribution of commune population",
    x = "Population per commune",
    y = "Number of communes"
  )
```

Expected interpretation:

-   The histogram is typically right-skewed.
-   Most communes have relatively smaller populations, with a few much larger communes.

## Exercise 4

**Task:** Compute z-scores of `total_population` and list communes with `z_population > 2`.

```{r}
#| label: csv-solution-4

pop_z <- pop |>
  mutate(
    total_population = FEMMES_MINEURES + HOMMES_MINEURS + FEMMES_MAJEURES + HOMMES_MAJEURS,
    z_population = as.numeric(scale(total_population))
  )

pop_z |>
  filter(z_population > 2) |>
  arrange(desc(z_population)) |>
  select(COMMUNE_NOM, total_population, z_population)
```

Why this works:

-   `scale()` standardizes values to mean 0 and standard deviation 1.
-   `z_population > 2` identifies unusually large communes relative to the full distribution.

## Exercise 5

**Task:** Plot `total_population` vs `adult_share` and describe the pattern.

```{r}
#| label: csv-solution-5
#| fig-width: 8
#| fig-height: 5

pop |>
  mutate(
    total_population = FEMMES_MINEURES + HOMMES_MINEURS + FEMMES_MAJEURES + HOMMES_MAJEURS,
    adult_share = (FEMMES_MAJEURES + HOMMES_MAJEURS) / total_population
  ) |>
  ggplot(aes(x = total_population, y = adult_share)) +
  geom_point(alpha = 0.7, color = "steelblue") +
  geom_smooth(method = "lm", se = FALSE, color = "firebrick") +
  labs(
    title = "Adult share vs commune population",
    x = "Total population",
    y = "Adult share"
  )
```

Interpretation note:

-   Use the regression line as a rough trend indicator.
-   In many datasets, adult share varies only moderately across commune size.

# Gutenberg + tidytext exercises — solutions

First build the baseline token table:

```{r}
#| label: text-base
#| message: false

base_books <- gutenberg_download(c(11, 1342), meta_fields = "title")
base_books <- base_books |>
  filter(
    is.na(lag(text)) | text != lag(text),
    .by = gutenberg_id
  )

base_tokens <- base_books |>
  unnest_tokens(word, text) |>
  anti_join(stop_words, by = join_by(word))
```

## Exercise 1

**Task:** Add one extra Gutenberg text and compute top 20 non-stopwords.

```{r}
#| label: text-solution-1

extra_book <- gutenberg_download(84, meta_fields = "title") # Frankenstein
extra_book <- extra_book |>
  filter(
    is.na(lag(text)) | text != lag(text),
    .by = gutenberg_id
  )

extra_top_words <- extra_book |>
  unnest_tokens(word, text) |>
  anti_join(stop_words, by = join_by(word)) |>
  count(title, word, sort = TRUE) |>
  slice_head(n = 20)

extra_top_words
```

Why this works:

-   `unnest_tokens()` creates one-token-per-row format.
-   `anti_join(stop_words, by = join_by(word))` removes common function words.
-   `count(sort = TRUE)` yields frequency ranking directly.

## Exercise 2

**Task:** Build a three-book dataset and compare top words.

```{r}
#| label: text-solution-2
#| fig-width: 10
#| fig-height: 6

extra_tokens <- extra_book |>
  unnest_tokens(word, text) |>
  anti_join(stop_words, by = join_by(word))

three_book_tokens <- bind_rows(base_tokens, extra_tokens)

three_top <- three_book_tokens |>
  count(title, word, sort = TRUE) |>
  group_by(title) |>
  slice_head(n = 12) |>
  ungroup()

three_top |>
  mutate(word = fct_reorder(word, n)) |>
  ggplot(aes(x = n, y = word, fill = title)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(vars(title), scales = "free_y") +
  labs(
    title = "Top words in three Gutenberg books",
    x = "Count",
    y = NULL
  )
```

Interpretation note:

-   Shared high-frequency content words may reflect common literary vocabulary.
-   Distinctive words often point to characters, setting, or genre themes.

## Exercise 3

**Task:** Compute tf-idf and interpret high tf-idf words.

```{r}
#| label: text-solution-3

three_tfidf <- three_book_tokens |>
  count(title, word, sort = TRUE) |>
  bind_tf_idf(word, title, n) |>
  arrange(desc(tf_idf))

three_tfidf |>
  group_by(title) |>
  slice_head(n = 10) |>
  ungroup() |>
  select(title, word, n, tf_idf)
```

Interpretation guideline:

-   High tf-idf means a word is frequent in one text but not frequent across all texts.
-   Character names and topic-specific words are common high tf-idf candidates.

## Exercise 4

**Task:** Compute bigrams for one selected book and list frequent non-stopword bigrams.

```{r}
#| label: text-solution-4

book_bigrams <- extra_book |>
  unnest_tokens(bigram, text, token = "ngrams", n = 2) |>
  separate_wider_delim(bigram, delim = " ", names = c("word1", "word2")) |>
  filter(!is.na(word1), !is.na(word2)) |>
  anti_join(stop_words, by = join_by(word1 == word)) |>
  anti_join(stop_words, by = join_by(word2 == word)) |>
  count(word1, word2, sort = TRUE)

book_bigrams |>
  mutate(bigram = str_c(word1, word2, sep = " ")) |>
  select(bigram, n) |>
  slice_head(n = 20)
```

Why this works:

-   `token = "ngrams", n = 2` produces adjacent two-word sequences.
-   Removing stop words from both positions highlights more semantic phrases.

## Exercise 5

**Task:** Compare lexical diversity across the three books.

```{r}
#| label: text-solution-5

three_book_tokens |>
  summarise(
    total_tokens = n(),
    unique_tokens = n_distinct(word),
    lexical_diversity = unique_tokens / total_tokens,
    .by = title
  ) |>
  arrange(desc(lexical_diversity))
```

Interpretation note:

-   Higher lexical diversity means a larger share of unique words.
-   The metric is sensitive to document length and genre.

# XML / LOD exercises — solutions

Build a reusable entry table:

```{r}
#| label: xml-base
#| message: false

lod_xml <- read_xml(resolve_path("new_lod-art.xml"))
entry_nodes <- xml_find_all(lod_xml, "//entry")
set.seed(42)
sample_n <- min(1000, length(entry_nodes))
entry_idx <- sample.int(length(entry_nodes), size = sample_n)
entry_sample <- entry_nodes[entry_idx]

node_text_or_na <- function(node, xpath) {
  candidate <- xml_find_first(node, xpath)
  if (inherits(candidate, "xml_missing")) {
    NA_character_
  } else {
    xml_text(candidate)
  }
}
```

For teaching speed, the solutions below use `entry_sample` (up to 1,000 random entries), not the full XML.

```{r}
#| label: xml-full-note
#| eval: false

# Full run (slow, often many minutes):
# entry_sample <- entry_nodes
```

## Exercise 1

**Task:** Extract `lemma`, `ipa`, and `n_examples`.

```{r}
#| label: xml-solution-1

xml_tbl <- entry_sample |>
  map(
    \(entry_node) {
      tibble(
        lemma = node_text_or_na(entry_node, ".//lemma"),
        ipa = node_text_or_na(entry_node, ".//ipa"),
        n_examples = xml_find_all(entry_node, ".//meaning/examples/example") |> length()
      )
    }
  ) |>
  list_rbind()

xml_tbl
```

Why this works:

-   Each XML `entry` becomes exactly one tibble row.
-   Counting `<example>` nodes captures usage richness per entry.

## Exercise 2

**Task:** Top 25 first English translations and discuss limits.

```{r}
#| label: xml-solution-2

xml_first_translation <- entry_sample |>
  map(
    \(entry_node) {
      en_translations <- xml_find_all(
        entry_node,
        ".//meaning/targetLanguage[@lang='en']/translation"
      ) |>
        xml_text()

      tibble(
        lemma = node_text_or_na(entry_node, ".//lemma"),
        first_en_translation = en_translations |> first(default = NA_character_)
      )
    }
  ) |>
  list_rbind()

xml_first_translation |>
  filter(!is.na(first_en_translation)) |>
  count(first_en_translation, sort = TRUE) |>
  slice_head(n = 25)
```

Limitation discussion:

-   Using only the first translation ignores polysemy and context.
-   Different lemmas may share one broad English gloss, reducing specificity.
-   XML ordering may be editorial rather than frequency-based.

## Exercise 3

**Task:** Sample 30 entries with translations and inspect semantic alignment.

```{r}
#| label: xml-solution-3

set.seed(42)

xml_first_translation |>
  filter(!is.na(first_en_translation), !is.na(lemma)) |>
  slice_sample(n = 30) |>
  arrange(lemma)
```

Interpretation guideline:

-   Check whether the English gloss is plausible for the lemma.
-   Mark ambiguous or overly broad glosses for manual follow-up.

## Exercise 4

**Task:** Summarise average, median, and maximum `n_examples`.

```{r}
#| label: xml-solution-4

xml_tbl |>
  summarise(
    mean_n_examples = mean(n_examples),
    median_n_examples = median(n_examples),
    max_n_examples = max(n_examples)
  )
```

Why this works:

-   `summarise()` provides compact descriptive statistics for usage examples.
-   Comparing mean and median helps detect skewness.

## Exercise 5

**Task:** Keep entries with both IPA and an English translation, then compute the share.

```{r}
#| label: xml-solution-5

xml_joined <- xml_tbl |>
  left_join(
    xml_first_translation |>
      select(lemma, first_en_translation),
    by = join_by(lemma)
  ) |>
  mutate(has_ipa_and_en = !is.na(ipa) & !is.na(first_en_translation))

xml_joined |>
  summarise(
    n_entries = n(),
    n_has_ipa_and_en = sum(has_ipa_and_en),
    share_has_ipa_and_en = n_has_ipa_and_en / n_entries
  )
```

Interpretation note:

-   This gives a simple coverage metric for entries that are useful in bilingual phonological work.

## Exercise 6

**Task:** Extract `partofspeech` with `lemma` and list the 20 most frequent labels.

```{r}
#| label: xml-solution-6

xml_pos <- entry_sample |>
  map(
    \(entry_node) {
      pos_values <- xml_find_all(entry_node, ".//partofspeech") |>
        xml_text()

      tibble(
        lemma = node_text_or_na(entry_node, ".//lemma"),
        partofspeech = pos_values |> first(default = NA_character_)
      )
    }
  ) |>
  list_rbind()

xml_pos |>
  filter(!is.na(partofspeech), partofspeech != "") |>
  count(partofspeech, sort = TRUE) |>
  slice_head(n = 20)
```

Why this works:

-   `xml_find_all(..., ".//partofspeech")` collects part-of-speech nodes inside each entry.
-   `first(default = NA_character_)` keeps one value per entry and handles missing tags.
-   `count(..., sort = TRUE)` gives a frequency profile of grammatical categories.
