---
title: "Session 4"
subtitle: "Language Analytics in R"
author: "Peter Gilles"
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    code-fold: false
editor: visual
---

# 1. Maps with ggplot2

## Setup

Run this chunk once at the beginning. Data files in the `sessions/` folder: `communes_df.RDS`, `centroids_label.RDS`, `butterfly.rds`. Population counts come from the [RNPP age pyramid CSV](https://download.data.public.lu/resources/registre-national-des-personnes-physiques-rnpp-pyramide-dage-par-commune-population-age-pyramid-per-municipality/20260407-111759/31-03-2026-rnrpp-pyramide-age-commune.csv) on data.public.lu.

```{r}
#| message: false

library(ggplot2)
library(dplyr)
library(tidyr)
library(stringr)
library(readr)
library(gridExtra)
```

## Maps with Luxembourg examples

For a lightweight first map, we use polygon data from `map_data("world")` and zoom into Luxembourg.

```{r}
#| label: ggplot-map-lux
#| fig-width: 11
#| fig-height: 7
#| out-width: "100%"

world_map <- map_data("world")

ggplot(world_map, aes(x = long, y = lat, group = group)) +
  geom_polygon(fill = "grey90", color = "purple", linewidth = 0.8) +
  coord_quickmap(xlim = c(4.5, 7.0), ylim = c(49.2, 50.4)) +
  labs(
    title = "Luxembourg in regional context",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_minimal()
```

In addition, we need the polygon data for the communes and geolocalisation data for the main cities per commune.

```{r}
#| label: ggplot-map-lux-data

world_map <- map_data("world")
communes_df <- readRDS("communes_df.RDS")
centroids_label <- readRDS("centroids_label.RDS")

lux_map <- world_map |>
  filter(region == "Luxembourg")

lux_map
```

**Step 1:** regional background.

```{r}
#| label: ggplot-map-lux-step-1
#| fig-width: 8
#| fig-height: 6
#| out-width: "100%"

p1 <- ggplot() +
  geom_polygon(
    data = world_map,
    aes(x = long, y = lat, group = group),
    fill = "grey92", color = "white", linewidth = 0.2
  ) +
  coord_quickmap(xlim = c(5.5, 6.7), ylim = c(49.2, 50.4)) +
  labs(
    title = "Step 1: Regional background",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_minimal()

p1
```

**Step 2:** add commune polygons. Note that we add now one ggplot object (p1, p2 ...) on top of each other.

```{r}
#| label: ggplot-map-lux-step-2
#| fig-width: 8
#| fig-height: 6
#| out-width: "100%"

p2 <- p1 +
  geom_polygon(
    data = communes_df,
    aes(x = long, y = lat, group = group),
    fill = "#fdebd0", color = "#d35400", linewidth = 0.15, alpha = 0.8
  ) +
  labs(title = "Step 2: + Commune polygons")

p2
```

**Step 3:** add Luxembourg country outline.

```{r}
#| label: ggplot-map-lux-step-3
#| fig-width: 8
#| fig-height: 6
#| out-width: "100%"

p3 <- p2 +
  geom_polygon(
    data = lux_map,
    aes(x = long, y = lat, group = group),
    fill = NA, color = "black", linewidth = 0.5
  ) +
  labs(title = "Step 3: + Country outline")

p3
```

**Step 4:** add centroid points for important places.

```{r}
#| label: ggplot-map-lux-step-4
#| fig-width: 8
#| fig-height: 6
#| out-width: "100%"

p4 <- p3 +
  geom_point(
    data = centroids_label,
    aes(x = long, y = lat),
    color = "navy", size = .8
  ) +
  labs(title = "Step 4: + Important place points")

p4
```

**Step 5:** add city labels (final map).

```{r}
#| label: ggplot-map-lux-step-5
#| fig-width: 8
#| fig-height: 6
#| out-width: "100%"

p5 <- p4 +
  geom_text(
    data = centroids_label,
    aes(x = long, y = lat, label = label),
    size = 1.8, nudge_y = 0.015, color = "navy"
  ) +
  labs(
    title = "Step 5: Final map",
    subtitle = "Communes (polygons) + centroid city labels",
    x = "Longitude",
    y = "Latitude"
  )

p5
```

You can also combine the five map states into one overview grid:

```{r}
#| label: ggplot-map-layers-grid
#| fig-width: 13
#| fig-height: 10
#| out-width: "100%"

final_map <- gridExtra::grid.arrange(p1, p2, p3, p4, p5, ncol = 2)
final_map

ggsave("map.png", plot = final_map, dpi = 600)
```

## Load RNPP population data

The RNPP file has one row per commune and sex (`F` / `M`), with population counts in 5-year age columns (`AGE00_05`, `AGE05_10`, …). We load it once and derive commune totals for the maps below.

```{r}
#| label: rnpp-load

pyramid_url <- "https://download.data.public.lu/resources/registre-national-des-personnes-physiques-rnpp-pyramide-dage-par-commune-population-age-pyramid-per-municipality/20260407-111759/31-03-2026-rnrpp-pyramide-age-commune.csv"

age_raw <- read_csv(pyramid_url)

# one row per commune: sum all age columns (both sexes)
pop_by_commune <- age_raw |>
  mutate(across(starts_with("AGE"), as.numeric)) |>
  summarise(
    total_population = sum(across(starts_with("AGE")), na.rm = TRUE),
    .by = COMMUNE_NOM
  )
```

## Population per commune (choropleth)

This map visualizes total population (`total_population`) by commune.

How it works:

1.  Aggregate RNPP age columns to one total per commune (`pop_by_commune`).
2.  Join population values to commune polygons via commune name (`id` vs `COMMUNE_NOM`).
3.  Use `fill = total_population` to encode population with a color gradient.
4.  Optionally add commune labels at centroids for orientation.

**Join.** `left_join()` attaches `total_population` to every polygon row. The join uses two different column names: polygon `id` (commune name) matches RNPP `COMMUNE_NOM`. Rows without a match keep `NA` for `total_population` (grey on the map).

```{r}
#| label: ggplot-map-population-communes
#| fig-width: 14
#| fig-height: 10
#| out-width: "100%"

communes_df <- readRDS("communes_df.RDS")
centroids_label <- readRDS("centroids_label.RDS")

communes_pop_map <- communes_df |>
  left_join(pop_by_commune, by = c("id" = "COMMUNE_NOM"))
```

**Plot.** In `aes()`, `group = group` is essential for polygons: each commune consists of many coordinate rows that must be drawn as one shape. `fill = total_population` maps population to colour. `geom_text()` uses a separate table (`centroids_label`); `inherit.aes = FALSE` prevents it from inheriting `fill` and `group` from the polygon layer. `coord_quickmap()` sets the map extent; `scale_fill_viridis_c()` applies a continuous colour scale with readable axis labels.

```{r}
#| label: ggplot-map-population-communes-plot
#| fig-width: 14
#| fig-height: 10
#| out-width: "100%"

ggplot(communes_pop_map, aes(x = long, y = lat, group = group, fill = log(total_population))) +
  geom_polygon(color = "white", linewidth = 0.15) +
  geom_text(
    data = centroids_label,
    aes(x = long, y = lat, label = label),
    inherit.aes = FALSE,
    size = 2.4,
    color = "grey20"
  ) +
  coord_quickmap(xlim = c(5.5, 6.7), ylim = c(49.2, 50.2)) +
  scale_fill_viridis_c(option = "plasma", labels = scales::label_number(big.mark = " ")) +
  labs(
    title = "Luxembourg: population by commune",
    subtitle = "RNPP, 31 March 2026 — sum of all age bands",
    fill = "Population",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_void()
```

- Brighter communes have higher population totals.
- Luxembourg City should stand out clearly as the most populous commune.
- A few communes were merged or renamed since the polygon file was built; unmatched polygons have no fill.

## Map exercise

### Exercise (population): choropleth with joins and aggregation

**Goal:** practice map joins and pre-aggregation for cartographic input.

Tasks:

1.  Join `communes_df` with `pop_by_commune` to map `total_population` by commune.
2.  Create the choropleth (`fill = total_population`) with commune labels.
3.  Build a 5-class quantile variable (`pop_quantile`) from `total_population`. Check the manual how to use `dplyr::ntile`.
4.  Plot a second choropleth using `fill = pop_quantile` (discrete classes).

::: {.callout-tip collapse="true"}
## Possible solution (click to expand)

```{r}
communes_df <- readRDS("communes_df.RDS")
centroids_label <- readRDS("centroids_label.RDS")

# join map polygons with RNPP commune totals
communes_pop_map <- communes_df |>
  left_join(pop_by_commune, by = c("id" = "COMMUNE_NOM"))

# continuous choropleth
ggplot(communes_pop_map, aes(x = long, y = lat, group = group, fill = total_population)) +
  geom_polygon(color = "white", linewidth = 0.15) +
  geom_text(
    data = centroids_label,
    aes(x = long, y = lat, label = label),
    inherit.aes = FALSE,
    size = 2.2
  ) +
  coord_quickmap(xlim = c(5.5, 6.7), ylim = c(49.2, 50.2)) +
  labs(title = "Population by commune", fill = "Population") +
  theme_minimal()

# aggregation into quantile groups
communes_pop_map <- communes_pop_map |>
  mutate(pop_quantile = dplyr::ntile(total_population, 5)) |>
  na.omit()

# discrete choropleth (5 classes)
ggplot(communes_pop_map, aes(x = long, y = lat, group = group, fill = factor(pop_quantile))) +
  geom_polygon(color = "white", linewidth = 0.15) +
  coord_quickmap(xlim = c(5.5, 6.7), ylim = c(49.2, 50.2)) +
  labs(title = "Population quantiles by commune", fill = "Quantile") +
  theme_minimal()
```
:::

## Maps by sex with `facet_wrap()`

The RNPP table has separate rows for `F` and `M`. After summing population per commune and sex, we can plot **two choropleths side by side** with `facet_wrap(~ SEXE)`.

How it works:

1.  Aggregate `total_population` by `COMMUNE_NOM` and `SEXE` (not merged across sexes).
2.  Join to `communes_df` — each commune polygon appears twice (once per sex).
3.  Add `facet_wrap(~ SEXE)` so ggplot draws one map panel per sex.

```{r}
#| label: rnpp-map-facet-sex
#| fig-width: 14
#| fig-height: 7
#| out-width: "100%"

pop_by_commune_sex <- age_raw |>
  mutate(across(starts_with("AGE"), as.numeric)) |>
  summarise(
    total_population = sum(across(starts_with("AGE")), na.rm = TRUE),
    .by = c(COMMUNE_NOM, SEXE)
  )

communes_pop_sex <- communes_df |>
  left_join(pop_by_commune_sex, by = c("id" = "COMMUNE_NOM")) |>
  na.omit()

ggplot(communes_pop_sex, aes(x = long, y = lat, group = group, fill = ntile(total_population, 5))) +
  geom_polygon(color = "white", linewidth = 0.15) +
  coord_quickmap(xlim = c(5.5, 6.7), ylim = c(49.2, 50.2)) +
  facet_wrap(
    ~ SEXE,
    labeller = labeller(SEXE = c(F = "Women", M = "Men"))
  ) +
  scale_fill_viridis_c(option = "plasma", labels = scales::label_number(big.mark = " ")) +
  labs(
    title = "Population by commune and sex",
    subtitle = "RNPP, 31 March 2026 — same fill scale in both panels",
    fill = "Population"
  ) +
  theme_void() +
  theme(strip.text = element_text(face = "bold"))
```

- **`facet_wrap(~ SEXE)`** splits the plot into one panel per sex. The formula `~ SEXE` means “create facets from the `SEXE` column”. Each panel reuses the same `geom_polygon()` call; ggplot filters the data automatically.
- **`labeller = labeller(...)`** renames facet strip labels (`F` → “Women”, `M` → “Men”) without changing the underlying data.
- **`scale_fill_viridis_c()`** defines one continuous colour scale for both panels, so darker/brighter communes are comparable across women and men.
- **`theme(strip.text = ...)`** formats the facet titles (the strips above each small map).

Both panels share the same colour scale by default, which makes the two maps comparable.

## Age pyramids

Using the same `age_raw` table, reshape the wide age columns into long format. For a classic pyramid, plot men with **negative** counts (left) and women with **positive** counts (right).

### One commune

```{r}
#| label: rnpp-age-pyramid-setup

commune_pick <- "Differdange"

age_levels <- age_raw |>
  select(starts_with("AGE")) |>
  names() |>
  str_remove("^AGE") |>
  str_replace("_", "-") |>
  (\(x) if_else(x == "100-", "100+", x))()
```

**Reshape.** The RNPP table is **wide**: one column per age band. `pivot_longer()` stacks them into rows (`age_band`, `count`). We then clean band labels (`AGE20_25` → `20-25`) and fix the order with `factor(..., levels = age_levels)`. Multiplying men's counts by −1 places their bars on the **left** of the pyramid; women stay positive on the **right**.

```{r}
#| label: rnpp-age-pyramid-data
#| fig-width: 10
#| fig-height: 8
#| out-width: "100%"

age_pyramid <- age_raw |>
  filter(COMMUNE_NOM == commune_pick) |>
  pivot_longer(starts_with("AGE"), names_to = "age_band", values_to = "count") |>
  mutate(
    age_band = str_remove(age_band, "^AGE"),
    age_band = str_replace(age_band, "_", "-"),
    age_band = if_else(age_band == "100-", "100+", age_band),
    age_band = factor(age_band, levels = age_levels),
    count = if_else(SEXE == "M", -count, count)
  )
```

**Plot.** `geom_col()` draws horizontal bars; `geom_vline(xintercept = 0)` marks the centre line between women and men. `scale_x_continuous(labels = ...)` uses `abs()` so axis labels show positive numbers even though men's values are negative.

```{r}
#| label: rnpp-age-pyramid-plot
#| fig-width: 10
#| fig-height: 8
#| out-width: "100%"

ggplot(age_pyramid, aes(x = count, y = age_band, fill = SEXE)) +
  geom_col(width = 0.85) +
  geom_vline(xintercept = 0, linewidth = 0.3) +
  scale_fill_manual(
    values = c(F = "#c0392b", M = "#2980b9"),
    labels = c(F = "Women", M = "Men")
  ) +
  scale_x_continuous(labels = \(x) format(abs(x), big.mark = " ")) +
  labs(
    title = paste("Age pyramid:", commune_pick),
    subtitle = "RNPP, 31 March 2026 — counts by 5-year age band",
    x = "Population",
    y = "Age group",
    fill = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

Change `commune_pick` (e.g. to `"Luxembourg"` or `"Esch-sur-Alzette"`) to compare age structures.

### Entire country

Aggregate `age_raw` across all communes, keeping all age columns and grouping by sex.

**Summarise.** `summarise(across(...), .by = SEXE)` adds up every age column **across all communes**, but keeps women and men in separate rows. The result is still wide (one column per age band).

**Age order.** The pipeline on `age_levels` extracts column names, removes the `AGE` prefix, and turns underscores into hyphens. The anonymous function `\(x) if_else(...)` is a short way to relabel the top band `100-` as `100+`.

**Long format.** The same `pivot_longer()` + `mutate()` steps as above turn the national table into pyramid-ready data.

```{r}
#| label: rnpp-age-pyramid-national
#| fig-width: 10
#| fig-height: 8
#| out-width: "100%"

age_total <- age_raw |>
  summarise(
    across(starts_with("AGE"), \(x) sum(x, na.rm = TRUE)),
    .by = SEXE
  )

age_levels <- age_total |>
  select(starts_with("AGE")) |>
  names() |>
  str_remove("^AGE") |>
  str_replace("_", "-") |>
  (\(x) if_else(x == "100-", "100+", x))()

age_pyramid <- age_total |>
  pivot_longer(starts_with("AGE"), names_to = "age_band", values_to = "count") |>
  mutate(
    age_band = str_remove(age_band, "^AGE"),
    age_band = str_replace(age_band, "_", "-"),
    age_band = if_else(age_band == "100-", "100+", age_band),
    age_band = factor(age_band, levels = age_levels),
    count = if_else(SEXE == "M", -count, count)
  )

ggplot(age_pyramid, aes(x = count, y = age_band, fill = SEXE)) +
  geom_col(width = 0.85) +
  geom_vline(xintercept = 0, linewidth = 0.3) +
  scale_fill_manual(
    values = c(F = "#c0392b", M = "#2980b9"),
    labels = c(F = "Women", M = "Men")
  ) +
  scale_x_continuous(labels = \(x) format(abs(x), big.mark = " ")) +
  labs(
    title = paste("Age pyramid: Luxembourg"),
    subtitle = "RNPP, 31 March 2026 — counts by 5-year age band",
    x = "Population",
    y = "Age group",
    fill = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

```

# 2. Data exploration

This section uses the **Schnessen** butterfly dataset (`butterfly.rds`): crowd-sourced recordings of the word for “butterfly” across Luxembourg, with sociolinguistic metadata and commune information.

## Load and inspect `butterfly.rds`

```{r}
#| label: butterfly-load

butterfly <- readRDS("butterfly.rds")

glimpse(butterfly)
```

Each row is one recording. Key variables for this analysis:

| Variable            | Meaning                                    |
|---------------------|--------------------------------------------|
| `Variant_Päiperlek` | Coded variant of the target word           |
| `Alter`             | Age group                                  |
| `Geschlecht`        | Gender                                     |
| `Ausbildung`        | Education level                            |
| `Dialektgebiet`     | Dialect area (north / east / south / west) |
| `Gemeng`            | Commune (factor, matches polygon `id`)     |

```{r}
#| label: butterfly-variant-overview

butterfly |>
  count(Variant_Päiperlek, sort = TRUE) |>
  mutate(share = n / sum(n))
```

`FALSE` means the target word was not produced (or could not be coded). For variant statistics we exclude these rows.

```{r}
#| label: butterfly-filter

butterfly_variants <- butterfly |>
  filter(Variant_Päiperlek != "FALSE")

nrow(butterfly_variants)
```

## Variant statistics by age, gender, education, and dialect area

Reshape the four sociolinguistic variables into long format, then compute counts and **row-wise shares** (distribution of variants within each age group, etc.).

**Convert to character.** `Alter`, `Geschlecht`, and the other columns are factors or ordered factors. `pivot_longer()` needs a common type in the `values` column, so we convert them to character first.

**Long format.** `pivot_longer()` stacks the four columns into two new ones: `variable` (which social factor) and `level` (the actual category, e.g. “25 bis 34”).

**Counts and shares.** `count()` tallies recordings per social group and variant. `group_by(variable, level)` + `mutate(share = n / sum(n))` computes the **proportion of each variant within** each age band (or gender, etc.) — the values in one row of the output sum to 1.

```{r}
#| label: butterfly-variant-stats

variant_by_social <- butterfly_variants |>
  mutate(across(
    c(Alter, Geschlecht, Ausbildung, Dialektgebiet),
    \(x) as.character(x)
  )) |>
  pivot_longer(
    c(Alter, Geschlecht, Ausbildung, Dialektgebiet),
    names_to = "variable",
    values_to = "level"
  ) |>
  count(variable, level, Variant_Päiperlek) |>
  group_by(variable, level) |>
  mutate(share = n / sum(n), .groups = "drop") |>
  arrange(variable, level, desc(n))

variant_by_social
```

**Age (`Alter`)** — share of each variant by age band:

```{r}
#| label: butterfly-stats-alter

variant_by_social |>
  filter(variable == "Alter") |>
  select(level, Variant_Päiperlek, n, share)
```

**Gender (`Geschlecht`):**

```{r}
#| label: butterfly-stats-geschlecht

variant_by_social |>
  filter(variable == "Geschlecht") |>
  select(level, Variant_Päiperlek, n, share)
```

**Education (`Ausbildung`):**

```{r}
#| label: butterfly-stats-ausbildung

variant_by_social |>
  filter(variable == "Ausbildung") |>
  select(level, Variant_Päiperlek, n, share)
```

**Dialect area (`Dialektgebiet`):**

```{r}
#| label: butterfly-stats-dialektgebiet

variant_by_social |>
  filter(variable == "Dialektgebiet") |>
  select(level, Variant_Päiperlek, n, share)
```

Bar chart: top variants only (at least 20 tokens overall) to keep the plot readable.

```{r}
#| label: butterfly-variant-barchart
#| fig-width: 12
#| fig-height: 8
#| out-width: "100%"

top_variants <- butterfly_variants |>
  count(Variant_Päiperlek, sort = TRUE) |>
  filter(n >= 20) |>
  pull(Variant_Päiperlek)

butterfly_variants |>
  filter(Variant_Päiperlek %in% top_variants) |>
  mutate(across(
    c(Alter, Geschlecht, Ausbildung, Dialektgebiet),
    \(x) as.character(x)
  )) |>
  pivot_longer(
    c(Alter, Geschlecht, Ausbildung, Dialektgebiet),
    names_to = "variable",
    values_to = "level"
  ) |>
  count(variable, level, Variant_Päiperlek) |>
  group_by(variable, level) |>
  mutate(share = n / sum(n), .groups = "drop") |>
  ggplot(aes(x = level, y = share, fill = Variant_Päiperlek)) +
  geom_col(position = "stack") +
  facet_wrap(~ variable, scales = "free_x") +
  labs(
    title = "Butterfly variants by sociolinguistic category",
    x = NULL,
    y = "Share within group",
    fill = "Variant"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Map variants by commune

Aggregate to commune level: number of recordings, share of the standard form `Päiperlek`, and the **dominant** variant (most frequent code in that commune).

```{r}
#| label: butterfly-commune-stats

commune_variants <- butterfly_variants |>
  mutate(Gemeng = as.character(Gemeng)) |>
  summarise(
    n = n(),
    share_päiperlek = mean(Variant_Päiperlek == "Päiperlek"),
    .by = Gemeng
  ) |>
  left_join(
    butterfly_variants |>
      mutate(Gemeng = as.character(Gemeng)) |>
      count(Gemeng, Variant_Päiperlek) |>
      slice_max(n, n = 1, by = Gemeng, with_ties = FALSE) |>
      select(Gemeng, dominant_variant = Variant_Päiperlek),
    by = "Gemeng"
  )

commune_variants
```

Join to commune polygons and plot the dominant variant:

```{r}
#| label: butterfly-map-dominant
#| fig-width: 14
#| fig-height: 10
#| out-width: "100%"

communes_df <- readRDS("communes_df.RDS")

communes_butterfly <- communes_df |>
  left_join(commune_variants, by = c("id" = "Gemeng"))

ggplot() +
  geom_polygon(data = communes_butterfly, aes(x = long, y = lat, group = group, fill = dominant_variant), color = "white", linewidth = 0.15) +
  geom_text(data = centroids_label,
    aes(x = long, y = lat, label = label),
    size = 2.8, nudge_y = 0.015, color = "navy"
  ) +
  coord_quickmap(xlim = c(5.5, 6.7), ylim = c(49.2, 50.2)) +
  scale_fill_brewer(palette = "Set3", na.value = "grey85") +
  labs(
    title = "Dominant butterfly variant by commune",
    subtitle = "Schnessen / butterfly.rds — communes without recordings in grey",
    fill = "Dominant variant"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

Continuous map: share of the variant `Päiperlek` per commune (among coded responses).

```{r}
#| label: butterfly-map-share
#| fig-width: 14
#| fig-height: 10
#| out-width: "100%"

ggplot(communes_butterfly, aes(x = long, y = lat, group = group, fill = share_päiperlek)) +
  geom_polygon(color = "white", linewidth = 0.15) +
  coord_quickmap(xlim = c(5.5, 6.7), ylim = c(49.2, 50.2)) +
  scale_fill_viridis_c(option = "mako", labels = scales::percent_format(), limits = c(0, 1)) +
  labs(
    title = "Share of variant “Päiperlek” by commune",
    subtitle = "Among coded Schnessen responses per commune",
    fill = "Share Päiperlek"
  ) +
  theme_minimal()
```
