---
title: "Session 1 — Introduction: R workflow and first steps"
subtitle: "Language Analytics in R"
author: "Peter Gilles"
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    code-fold: false
  revealjs:
    output-file: session-01-slides.html
    slide-level: 2
    smaller: true
    incremental: true
    transition: slide
    scrollable: true
    include-in-header:
      - text: |
          <style>
          /* Scrollable slides: constrain height so overflow can scroll (see SO 76962902) */
          .reveal .slide.scrollable {
            height: 100% !important;
            max-height: 100vh !important;
            overflow-y: auto !important;
            overflow-x: hidden;
            box-sizing: border-box;
          }
          </style>
editor: visual
---

## Learning objectives {.scrollable}

By the end of this session you will:

-   Use RStudio and run R code in the console and in a Quarto document.
-   Create and inspect objects with the assignment operator (`<-`).
-   Write clear names and comments, and apply basic code style.
-   Call functions with correct syntax and use the pipe (`|>`).
-   Run and render a minimal Quarto (.qmd) script.

This course focuses on **linguistic and language data in R**. Examples and exercises use words, frequencies, and simple text data so you can quickly apply the same ideas to your own corpora and experiments.

The material draws on [R for Data Science (2e)](https://r4ds.hadley.nz/) (chapters 2–4), [R for Linguists](https://eleanorchodroff.com/r-for-linguists/), and other linguistics-oriented R resources (see [Further reading](#further-reading)).

# RStudio and running R code

## The RStudio interface

-   **Console**: type R code and see results.
-   **Editor**: write and save scripts (e.g. `.qmd` or `.R`).
-   **Environment**: view objects you have created (datasets, variables).
-   **Files / Plots / Help**: navigate files, view plots, read help.

Run code from the editor with **Cmd+Enter** (Mac) or **Ctrl+Enter** (Windows/Linux).

## Getting help

Prefix a command with `?`, e.g. `?mean` to call the help page for this command. Or use the `Help` pane in RStudio.

```{r}
# Access help for a function by name:
?mean

# Alternatively, use the help() function:
help("mean")

# Search help topics with apropos() or help.search():
apropos("lin")  # Shows all objects/functions with "lin" in the name
help.search("linear model")  # Looks up help topics related to linear models

# View example usage for functions:
example(mean)

# See the vignette (long-form documentation) for a package:
# vignette("dplyr")
```

There are numerous R [cheatsheets](https://posit.co/resources/cheatsheets/) that summarise command usages in a structured way.

## Packages

Install a package once with `install.packages("package_name")` or use the `packages` tab in RStudio. Load it every time you start R or open a document with `library(package_name)`:

```{r}
#| label: packages
#| message: false

# Core packages for data import and manipulation
library(tidyverse)
```

## 

## Coding basics {#sec-basics}

You can use R as a calculator (e.g. for reaction times, accuracy, or counts):

```{r}
#| label: calc-basics

1 / 200 * 30
(59 + 73 + 2) / 3
sin(pi / 2)
```

Create objects with **assignment**:

```{r}
#| label: assignment

calc <- (59 + 73 + 2) / 3

# e.g. store a value like VOT (ms) or word length
vot_ms <- 25
word_length <- nchar("linguistics")
```

Every assignment has the form:

``` text
object_name <- value
```

Read it as: *“object name gets value”*.

**Tip:** In RStudio use **Alt + -** (minus) to insert `<-` with spaces.

The value is stored but not printed. Type the object name to inspect it:

```{r}
vot_ms
word_length
```

Combine values into a **vector** with `c()` — like a column of observations (e.g. reaction times, word lengths, or a list of words):

```{r}
#| label: vectors-c
# numeric vector: e.g. reaction times in ms
rt_ms <- c(320, 415, 380, 290, 410)

# character vector: e.g. word forms or conditions
words <- c("cat", "dog", "mouse")

# a rose is a rose — last expression is printed in the report
stein <- c("a", "rose", "is", "a", "rose", "is", "a", "rose")
print(stein)
```

Instead of referencing every token in a vector individually, we can convert them to **factor**s.

```{r}
stein.fac <- as.factor(c("a", "rose", "is", "a", "rose", "is", "a", "rose"))
print(stein.fac)
```

Print as table

```{r}
table(stein.fac)
```

# Tibbles and data frames {.scrollable}

Vectors and factors hold a single column of values. For **tabular data** (rows and columns), R uses **data frames**. A **tibble** is a modern, tidyverse-style data frame: it prints nicely, doesn’t convert strings to factors by default, and works well with the pipe and dplyr.

## Creating a tibble

Use **`tibble()`** to build a table from named vectors (each vector becomes a column). Columns can be referred to by name in later code.

```{r}
#| label: tibble-intro

library(tibble)

# Example: a small table (e.g. word, frequency, part of speech)
lex <- tibble(
  word = c("the", "cat", "sat", "on", "mat"),
  freq = c(100, 5, 3, 50, 2),
  pos  = c("det", "noun", "verb", "prep", "noun")
)
lex
```

With **`data.frame()`** you get a classic R data frame; tibbles are usually preferred in the tidyverse because of their print format and consistent behaviour.

## Inspecting structure

-   **Print:** typing the name shows the first rows and column types (tibbles don’t flood the console).
-   **`glimpse()`** (dplyr): one row per column, with type and a few values.
-   **`nrow()`, `ncol()`, `names()`**: number of rows, number of columns, column names.

```{r}
#| label: tibble-inspect

glimpse(lex)
nrow(lex)
names(lex)
```

## Accessing columns

Use **`$`** to get one column as a vector: `lex$word`, `lex$freq`. You can use that in expressions (e.g. `mean(lex$freq)`) or pass it to functions. Later you’ll use dplyr verbs (`filter`, `select`, `mutate`) to work with whole tables without pulling columns out by hand.

```{r}
lex$word
mean(lex$freq)
```

# Comments and names

## Comments {.scrollable}

R ignores everything after `#` on a line. Use comments to explain *why* you did something (e.g. why you chose a threshold or excluded a condition), not just *what* the code does.

```{r}
#| label: comments

# word list for lexical decision practice
words <- c("cat", "dog", "mouse")

# mean reaction time (ms)
mean(rt_ms)
```

## Naming conventions {#sec-names .scrollable}

-   Names must **start with a letter** and contain only letters, numbers, `_`, and `.`.
-   Use **snake_case**: lowercase words separated by `_` (e.g. `word_list`, `response_time`).

```{r}
#| label: naming

response_time_ms <- 320
word_list <- c("the", "cat", "sat")
# Avoid: responseTimeMs, Word.List
```

R is case-sensitive and does not fix typos: `word_list` and `Word_list` are different objects.

**Tip:** Type a prefix and press **Tab** for autocomplete; use **↑** or **Cmd+↑** to recall previous commands.

# Calling functions

Functions are called like this:

``` text
function_name(argument1 = value1, argument2 = value2, ...)
```

Example: `seq()` creates sequences (e.g. for item IDs or conditions):

```{r}
#| label: seq-call

seq(from = 1, to = 10)
seq(1, 5, by = 0.5)
```

You can omit argument names for the first arguments. Use **Tab** after typing a function name to see arguments and help.

**Strings** (text) need **matched quotation marks** — essential when working with words and corpora:

```{r}
# quotation marks around strings
word <- "hello"

# Compare strings with == (e.g. for accuracy coding)
"linguistics" == "linguistics"
"linguistics" == "lingiustics"
```

If you see a `+` in the console, R is waiting for more input (often a missing `"` or `)`). Press **Escape** to cancel.

# Code style {#sec-style}

Good style makes code easier to read for you and others. We follow the [tidyverse style guide](https://style.tidyverse.org/).

## Names (reminder) {.scrollable}

-   Use lowercase, numbers, and `_`.
-   Prefer **long, descriptive names** (e.g. `participant_response_time` over `prt`).

## Spaces {.scrollable}

-   Put spaces around `+`, `-`, `==`, `<`, etc. (except `^`).
-   Put spaces around `<-`.
-   Put a space after every comma.

```{r}
#| label: style-spaces

# Strive for: spaces around operators and after comma
n_char <- nchar("word")
mean(rt_ms, na.rm = TRUE)
# Avoid: n_char<-nchar("word") or mean(rt_ms,na.rm=TRUE)
```

## Pipes {.scrollable}

The pipe `|>` (alternatively `%>%` when using package `magrittr`) passes the result on the left into the function on the right. Read it as *“then”*.

-   Put a space before `|>` and put `|>` at the **end** of the line.
-   Indent the next line by two spaces.

```{r}
c("Méindeg", "Dënschdeg", "Mëttwoch") |>
  nchar()
```

We’ll use the pipe with **language data** in the next section.

## Sectioning comments {.scrollable}

Break long scripts into sections:

``` r
# Load data --------------------------------------

# Plot data --------------------------------------
```

In RStudio: **Cmd+Shift+R** (Mac) or **Ctrl+Shift+R** (Windows/Linux) inserts a section header.

**Tip:** Install the **styler** package and use the command palette (**Cmd+Shift+P**) to format code automatically.

# Scripting in Quarto

## Why Quarto? {.scrollable}

-   Combine **text, code, and output** in one document — ideal for reporting on corpus or experiment results.
-   **Reproducible**: re-run and re-render to update tables and figures.
-   Output to HTML, PDF, or other formats.

## Running code in a .qmd file {.scrollable}

-   Create an empty Quarto file in RStudio: File \> New File \> Quarto Document ...
-   **Code chunks** are enclosed in ```` ```{r} ```` and ```` ``` ````.
-   Run one chunk: **Cmd+Enter** (Mac) or **Ctrl+Enter** (Windows/Linux).
-   Render the whole document: **Cmd+Shift+K** (or **Ctrl+Shift+K**) or click *Render*.

Example chunk options:

``` r
#| label: my-chunk
#| echo: true
#| eval: true
```

-   `label`: unique name for the chunk.
-   `echo`: whether to show the code in the output.
-   `eval`: whether to run the code.

# 
